---
title: "Binomial Logistic Regression for Colorectal Cancer Microbiome Classification"
author: "Harshini Chintala"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    df_print: paged
    fig_caption: true
    self_contained: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "##",
  fig.width = 7.2,
  fig.height = 5.2,
  dpi = 300,
  fig.align = "center",
  out.width = "92%",
  results = "markup"
)

options(width = 90, scipen = 999)
source("../../R/project_setup.R")
set.seed(1910)

# Classical notebook output locations.
project_dir <- classical_dir
results_dir <- file.path(classical_dir, "results")
models_dir <- file.path(classical_dir, "models")
data_dir <- file.path(classical_dir, "data")
reports_dir <- file.path(classical_dir, "reports")
figures_dir <- file.path(classical_dir, "figures")
for (d in c(results_dir, models_dir, data_dir, reports_dir, figures_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

```

## Notebook roadmap

**Purpose:** This notebook is one reproducible component of the CRC microbiome classification pipeline.

**Execution order**
1. Load the project environment and fixed random seed (`1910`).
2. Load only the validated project artifacts required for this analysis.
3. Perform the analysis exactly as documented below.
4. Generate tables/figures and save machine-readable outputs.
5. Record the session information used for the run.

**Output contract:** generated tables, figures, model objects, audit files, and session information are written under `02 Classical ML/results/`. The notebook itself renders to HTML with a floating table of contents, so the sections can be navigated from the left-hand panel in the knitted document.

**Reproducibility rule:** this repository preserves the validated analysis logic and project-wide seed. Environment restoration is handled by `renv.lock` for R and `requirements.txt` for Python. The notebooks do not silently install unpinned packages during knitting.

### Reproducibility constants

- Project seed: **1910**
- Discovery cohort: **ZellerG_2014**
- Primary task: **Healthy vs CRC; adenoma excluded**
- Frozen discovery split: **92 training / 22 held-out test samples**
- Primary feature panel: **20 Boruta-selected features**
- External cohorts: **FengQ_2015** and **VogtmannE_2016**

# Load Dataset


```{r}
# ==========================================================
# Load Dataset
# ==========================================================

library(curatedMetagenomicData)
library(TreeSummarizedExperiment)
library(tidyverse)
library(dplyr)
library(caret)
library(Boruta)
library(pROC)

# Download/load Zeller dataset
zeller <- curatedMetagenomicData(
  "ZellerG_2014.relative_abundance",
  dryrun = FALSE
)[[1]]

# Extract abundance matrix
abundance <- assay(zeller)

# Transpose:
# Rows = samples
# Columns = microbial species
X <- as.data.frame(t(abundance))

# Disease labels
y <- colData(zeller)$study_condition

# Combine into one dataframe
crc_data <- X
crc_data$disease <- y

# Check dimensions
dim(crc_data)

# Class distribution
table(crc_data$disease)
```

# Create Binary Dataset

```{r}
# 2. Create Binary Dataset
# ========================
binary_data <- crc_data %>%
  # 1. Rename "control" to "healthy" before filtering
  mutate(disease = case_match(disease, "control" ~ "healthy", .default = disease)) %>% 
  # 2. Filter down to your desired classes
  filter(disease %in% c("healthy", "CRC"))

# 3. Clean up the factor levels so "adenoma" is removed completely
binary_data$disease <- factor(binary_data$disease)

table(binary_data$disease)

```


# Train-Test Split

```{r}
# ==========================================================
# Shared Train-Test Split
# ==========================================================

train_index_path <- file.path(
  models_dir,
  "BinaryClass_Train_Index.rds"
)

if (file.exists(train_index_path)) {

  train_index <- readRDS(train_index_path)

} else {

  set.seed(1910)

  train_index <- createDataPartition(
    binary_data$disease,
    p = 0.80,
    list = FALSE
  )

  saveRDS(
    train_index,
    train_index_path
  )
}

train_data <- binary_data[train_index, ]
test_data  <- binary_data[-train_index, ]

cat("Training samples :", nrow(train_data), "\n")
cat("Testing samples  :", nrow(test_data), "\n")

table(train_data$disease)
table(test_data$disease)


```


```{r}
#==========================================================
# Prepare Predictor Matrix
#==========================================================

X_train_raw <- train_data %>%
  dplyr::select(-disease)

X_test_raw <- test_data %>%
  dplyr::select(-disease)

y_train <- train_data$disease
y_test <- test_data$disease
```

# Near-Zero Variance Filtering

```{r}
#==========================================================
# Near-Zero Variance Filtering
#==========================================================

# Detect near-zero variance features ONLY on training data
nzv <- nearZeroVar(X_train_raw)

cat("Near-zero variance features removed:", length(nzv), "\n")

# Remove the same columns from train and test
X_train_nzv <- X_train_raw[, -nzv, drop = FALSE]
X_test_nzv  <- X_test_raw[, -nzv, drop = FALSE]

cat("Remaining features:", ncol(X_train_nzv), "\n")
```

# Feature Selection using Boruta

```{r}
#---------------------------------------------------------------
# Feature Selection using Boruta
#---------------------------------------------------------------

set.seed(1910)

boruta_data <- data.frame(
  Disease = y_train,
  X_train_nzv,
  check.names = FALSE
)

boruta_model <- Boruta(
  Disease ~ .,
  data = boruta_data,
  doTrace = 2
)

boruta_final <- TentativeRoughFix(boruta_model)

selected_features <- getSelectedAttributes(
  boruta_final,
  withTentative = FALSE
)

# ==========================================================
# Save Canonical Shared Binary Feature List
# ==========================================================

canonical_feature_list <- data.frame(
  Feature = selected_features,
  stringsAsFactors = FALSE
)

write.csv(
  canonical_feature_list,
  file.path(
    results_dir,
    "Binary_Selected_20_Features.csv"
  ),
  row.names = FALSE
)

cat(
  "Canonical shared binary feature list saved successfully.\n"
)

cat("Number of selected features:",
    length(selected_features), "\n")

stopifnot(length(selected_features) == 20)

print(selected_features)

X_train_boruta <- X_train_nzv[, selected_features, drop = FALSE]
X_test_boruta  <- X_test_nzv[, selected_features, drop = FALSE]
```

```{r}
saveRDS(
  boruta_model,
  file.path(
    models_dir,
    "Boruta_Model.rds"
  )
)

cat("Boruta model saved successfully.\n")
```


# Feature Scaling

```{r}
# ==========================================================
# Feature Scaling
# ==========================================================

# Learn scaling parameters ONLY from training data
scaler <- preProcess(
  X_train_boruta,
  method = c("center", "scale")
)

X_train_scaled <- predict(scaler, X_train_boruta)
X_test_scaled  <- predict(scaler, X_test_boruta)

ncol(X_test_scaled)

head(colnames(X_test_scaled))

anyDuplicated(colnames(X_test_scaled))

cat("Training data scaled successfully.\n")
cat("Testing data scaled successfully.\n")
```


```{r}
saveRDS(
  scaler,
  file.path(
    models_dir,
    "Scaler.rds"
  )
)

cat("Scaler saved successfully.\n")
```

```{r}
#===============================================================
# Standardize Feature Names
#===============================================================

colnames(X_train_scaled) <- make.names(colnames(X_train_scaled))
colnames(X_test_scaled)  <- make.names(colnames(X_test_scaled))

cat("Feature names standardized successfully.\n")
```


# Repeated Stratified Cross-Validation

```{r}

# ===============================================================
# Shared Repeated Stratified Cross-Validation
# ===============================================================

cv_folds_path <- file.path(
  models_dir,
  "BinaryClass_CV_Folds.rds"
)

if (file.exists(cv_folds_path)) {

  cv_folds <- readRDS(cv_folds_path)

} else {

  set.seed(1910)

  cv_folds <- createMultiFolds(
    y_train,
    k = 10,
    times = 5
  )

  saveRDS(
    cv_folds,
    cv_folds_path
  )
}

shared_cv_control <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  index = cv_folds,
  classProbs = TRUE,
  savePredictions = "final",
  summaryFunction = twoClassSummary
)

```

```{r}
#===============================================================
# Train Binary Logistic Regression
#===============================================================

train_data <- data.frame(
  Disease = y_train,
  X_train_scaled,
  check.names = FALSE
)
set.seed(1910)

logistic_model <- caret::train(
  Disease ~ .,
  data = train_data,
  method = "glm",
  family = binomial(),
  metric = "ROC",
  trControl = shared_cv_control
)

logistic_model
```


```{r}
#===============================================================
# Prepare Test Data
#===============================================================

test_model_data <- data.frame(
  X_test_scaled,
  check.names = FALSE
)
```

# Cross-Validation Results

```{r}
#===============================================================
# Cross-Validation Results
#===============================================================

print(logistic_model)

logistic_model$results

cv_results <- logistic_model$results

write.csv(
    cv_results,
    file.path(
        results_dir,
        "BinomialLR_CrossValidation.csv"
    ),
    row.names = FALSE
)
```


# Variance Inflation Factor (VIF)

```{r}
#===============================================================
# Variance Inflation Factor (VIF)
#===============================================================

vif_values <- car::vif(logistic_model$finalModel)


vif_table <- data.frame(
  Feature = names(vif_values),
  VIF = vif_values
)

vif_table

write.csv(
  vif_table,
  file.path(
    results_dir,
    "BinomialLR_VIF.csv"
  ),
  row.names = FALSE
)
```

# Predict Probabilities

```{r}
#===============================================================
# Predict Probabilities
#===============================================================

prob_predictions <- predict(
  logistic_model,
  newdata = X_test_scaled,
  type = "prob"
)

head(prob_predictions)

summary(prob_predictions)
```


# Convert Probabilities to Predicted Classes

```{r}
#===============================================================
# Predict Classes
#===============================================================

predicted_class <- predict(
  logistic_model,
  newdata = X_test_scaled,
  type = "raw"
)

head(predicted_class)

table(predicted_class)
```

```{r}
#===============================================================
# Save Predictions
#===============================================================

prediction_results <- data.frame(
  SampleID = rownames(test_data),
  Actual = y_test,
  Predicted = predicted_class,
  Probability_CRC = prob_predictions$CRC,
  Probability_Healthy = prob_predictions$healthy,
  row.names = NULL
)
head(prediction_results)

write.csv(
  prediction_results,
  file.path(
    results_dir,
    "BinomialLR_Predictions.csv"
  ),
  row.names = FALSE
)

cat("Predictions saved successfully.\n")
```

# Confusion Matrix

```{r}
#===============================================================
# Confusion Matrix
#===============================================================

confusion_matrix <- confusionMatrix(
  predicted_class,
  y_test,
  positive = "CRC"
)

confusion_matrix
```

```{r}
#===============================================================
# Save Confusion Matrix
#===============================================================

capture.output(
  confusion_matrix,
  file = file.path(
    results_dir,
    "BinomialLR_ConfusionMatrix.txt"
  )
)

cat("Confusion matrix saved successfully.\n")
```

# Confusion Matrix Heatmap

```{r}
#===============================================================
# Confusion Matrix Heatmap
#===============================================================

confusion_df <- as.data.frame(confusion_matrix$table)

ggplot(confusion_df, aes(
  x = Prediction,
  y = Reference,
  fill = Freq
)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 6) +
  scale_fill_gradient(
    low = "white",
    high = "steelblue"
  ) +
  labs(
    title = "Binary Logistic Regression Confusion Matrix",
    x = "Predicted Class",
    y = "Actual Class"
  ) +
  theme_minimal()

ggsave(
  filename = file.path(
    figures_dir,
    "BinomialLR_ConfusionMatrixHeatmap.png"
  ),
  width = 6,
  height = 5,
  dpi = 300
)
```

# Overall Metrics

```{r}
#===============================================================
# Overall Evaluation Metrics
#===============================================================

overall_metrics <- data.frame(
  Accuracy = confusion_matrix$overall["Accuracy"],
  Kappa = confusion_matrix$overall["Kappa"],
  AccuracyLower = confusion_matrix$overall["AccuracyLower"],
  AccuracyUpper = confusion_matrix$overall["AccuracyUpper"],
  AccuracyPValue = confusion_matrix$overall["AccuracyPValue"],
  McNemarPValue = confusion_matrix$overall["McnemarPValue"]
)

overall_metrics

write.csv(
  overall_metrics,
  file.path(
    results_dir,
    "BinomialLR_OverallMetrics.csv"
  ),
  row.names = FALSE
)

cat("Overall metrics saved successfully.\n")
```

# Class-wise Metrics


```{r}
#===============================================================
# Class-wise Evaluation Metrics
#===============================================================

class_metrics <- as.data.frame(
  confusion_matrix$byClass
)

class_metrics$Metric <- rownames(class_metrics)

rownames(class_metrics) <- NULL

class_metrics

write.csv(
  class_metrics,
  file.path(
    results_dir,
    "BinomialLR_ClassMetrics.csv"
  ),
  row.names = FALSE
)

cat("Class-wise metrics saved successfully.\n")
```

# Matthews Correlation Coefficient (MCC)

```{r}
#===============================================================
# Matthews Correlation Coefficient (MCC)
#===============================================================

mcc_data <- data.frame(
  truth = y_test,
  prediction = predicted_class
)

mcc_value <- yardstick::mcc(
  mcc_data,
  truth = truth,
  estimate = prediction
)

mcc_value
```

```{r}
write.csv(
  mcc_value,
  file.path(
    results_dir,
    "BinomialLR_MCC.csv"
  ),
  row.names = FALSE
)

cat("MCC saved successfully.\n")
```


# ROC Curve

```{r}
#===============================================================
# ROC Curve
#===============================================================

roc_curve <- roc(
  response = y_test,
  predictor = prob_predictions$CRC,
  levels = c("healthy", "CRC")
)
plot(
  roc_curve,
  col = "blue",
  lwd = 2,
  main = "Binary Logistic Regression ROC Curve"
)

abline(
  a = 0,
  b = 1,
  lty = 2,
  col = "red"
)
```

```{r}

png(file.path(figures_dir, "ROC_Curve.png"), width = 15, height = 10, units = "in", res = 300)

plot(roc_curve) 

dev.off()

```

# AUC

```{r}
#===============================================================
# AUC
#===============================================================
auc_value <- auc(roc_curve)

auc_results <- data.frame(
  AUC = as.numeric(auc_value)
)

write.csv(
  auc_results,
  file.path(
    results_dir,
    "BinomialLR_AUC.csv"
  ),
  row.names = FALSE
)

cat("AUC saved successfully.\n")
```


# ROC Coordinates

```{r}
#===============================================================
# ROC Coordinates
#===============================================================
roc_coordinates <- data.frame(
  Specificity = roc_curve$specificities,
  Sensitivity = roc_curve$sensitivities,
  Threshold = roc_curve$thresholds
)

write.csv(
  roc_coordinates,
  file.path(
    results_dir,
    "BinomialLR_ROC_Coordinates.csv"
  ),
  row.names = FALSE
)

cat("ROC coordinates saved successfully.\n")
```


# Logistic Regression Coefficients

```{r}
#===============================================================
# Logistic Regression Coefficients
#===============================================================

coefficients_table <- summary(logistic_model)$coefficients

coefficients_table
```

```{r}
#===============================================================
# Extract Model Coefficients
#===============================================================

coefficients_df <- data.frame(
  Feature = names(coef(logistic_model$finalModel)),
  Estimate = coef(logistic_model$finalModel)
)

# Remove intercept
coefficients_df <- coefficients_df[
  coefficients_df$Feature != "(Intercept)",
]

coefficients_df
```

```{r}
write.csv(
  coefficients_df,
  file.path(
    results_dir,
    "BinomialLR_Coefficients.csv"
  ),
  row.names = FALSE
)

cat("Coefficient table saved successfully.\n")
```

# Calculate Odds Ratios

```{r}
#===============================================================
# Calculate Odds Ratios
#===============================================================

coefficients_df$OddsRatio <- exp(coefficients_df$Estimate)

coefficients_df
```

# Sort by Absolute Effect

```{r}
#===============================================================
# Sort Features by Effect Size
#===============================================================

coefficients_df <- coefficients_df[
  order(abs(coefficients_df$Estimate),
        decreasing = TRUE),
]

head(coefficients_df, 20)
```


# Logistic Regression Coefficient Plot


```{r}

# ================================================================
# Logistic Regression Coefficients and Top Biomarkers
# ================================================================

# Extract coefficients from the final fitted logistic model
coefficients_df <- data.frame(
  Feature = names(coef(logistic_model$finalModel)),
  Estimate = as.numeric(coef(logistic_model$finalModel)),
  stringsAsFactors = FALSE
)

# Remove intercept
coefficients_df <- coefficients_df[
  coefficients_df$Feature != "(Intercept)",
  ,
  drop = FALSE
]

# Calculate odds ratios
coefficients_df$OddsRatio <- exp(coefficients_df$Estimate)

# ================================================================
# Extract Genus + Species names
# ================================================================

extract_microbe_name <- function(x) {

  genus <- stringr::str_match(
    x,
    "g__([^.|]+)"
  )[, 2]

  species <- stringr::str_match(
    x,
    "s__([^.|]+)"
  )[, 2]

  if (
    !is.na(genus) &&
    !is.na(species)
  ) {
    return(
      paste(
        genus,
        species
      )
    )
  }

  # Fallback to original feature name
  return(x)
}

coefficients_df$Microbe <- vapply(
  coefficients_df$Feature,
  extract_microbe_name,
  character(1)
)

# ================================================================
# Rank features by absolute coefficient magnitude
# ================================================================

coefficients_df <- coefficients_df[
  order(
    abs(coefficients_df$Estimate),
    decreasing = TRUE
  ),
  ,
  drop = FALSE
]

coefficients_df$Rank <- seq_len(
  nrow(coefficients_df)
)

# ================================================================
# Save complete coefficient table
# ================================================================

write.csv(
  coefficients_df,
  file.path(
    results_dir,
    "BinomialLR_Coefficients.csv"
  ),
  row.names = FALSE
)

# ================================================================
# Extract Top 20 Biomarkers
# ================================================================

top20_biomarkers <- head(
  coefficients_df,
  20
)

top20_biomarkers$Direction <- ifelse(
  top20_biomarkers$Estimate > 0,
  "Positive",
  "Negative"
)

# Save dedicated Top-20 biomarker table
write.csv(
  top20_biomarkers,
  file.path(
    results_dir,
    "BinomialLR_Top20Biomarkers.csv"
  ),
  row.names = FALSE
)

# ================================================================
# Create Top-20 coefficient plot
# ================================================================

coefficient_plot <- ggplot(
  top20_biomarkers,
  aes(
    x = reorder(
      Microbe,
      abs(Estimate)
    ),
    y = Estimate,
    fill = Direction
  )
) +

  geom_col(
    width = 0.70
  ) +

  coord_flip() +

  labs(
    title = "Top 20 Logistic Regression Coefficients",
    subtitle = "Ranked by absolute coefficient magnitude",
    x = "Microbial taxon",
    y = "Logistic regression coefficient"
  ) +

  theme_minimal(
    base_size = 12
  ) +

  theme(
    legend.title = element_blank(),
    axis.text.y = element_text(
      size = 9
    )
  )

# Display in knitted HTML
print(coefficient_plot)

# ================================================================
# Save publication-quality figure
# ================================================================

ggsave(
  filename = file.path(
    figures_dir,
    "BinomialLR_Coefficients.png"
  ),
  plot = coefficient_plot,
  width = 9,
  height = 7,
  dpi = 300
)

cat(
  "Logistic regression coefficient table, Top-20 biomarker table, ",
  "and coefficient plot saved successfully.\n"
)
```



# Save Trained Model

```{r}
#===============================================================
# Save Trained Model
#===============================================================

saveRDS(
  logistic_model,
  file.path(
    models_dir,
    "BinomialLR_Model.rds"
  )
)
file.exists(
  file.path(
    models_dir,
    "BinomialLR_Model.rds"
  )
)

cat("Model saved successfully.\n")
```


# Session Information

```{r}
#===============================================================
# Session Information
#===============================================================

writeLines(
  capture.output(sessionInfo()),
  file.path(
    reports_dir,
    "BinomialLR_SessionInfo.txt"
  )
)

sessionInfo()
```

## Session information

The final section records the exact R session used for this knit. The same information is also saved as a text artifact so the HTML report and repository contain both a human-readable and machine-readable provenance record.

```{r session-information}
session_file <- file.path(
  results_dir,
  "session_info",
  "01_Binomial_Logistic_Regression_sessionInfo.txt"
)
write_session_info(session_file)
cat("Saved session information to:", session_file, "\n\n")
sessionInfo()
```
