---
title: "CLR Compositional Sensitivity Analysis"
subtitle: "Binary CRC vs Healthy Classification Using the Frozen 20-Feature Discovery Panel"
author: "Harshini Chintala"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    df_print: paged
    fig_caption: true
    self_contained: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "##",
  fig.width = 7.2,
  fig.height = 5.2,
  dpi = 300,
  fig.align = "center",
  out.width = "92%",
  results = "markup"
)

options(width = 90, scipen = 999)
source("../../R/project_setup.R")
set.seed(1910)

# Classical notebook output locations.
project_dir <- classical_dir
results_dir <- file.path(classical_dir, "results")
models_dir <- file.path(classical_dir, "models")
data_dir <- file.path(classical_dir, "data")
reports_dir <- file.path(classical_dir, "reports")
figures_dir <- file.path(classical_dir, "figures")
clr_results_dir <- file.path(
  results_dir,
  "clr_compositional_sensitivity"
)

clr_models_dir <- file.path(
  models_dir,
  "clr_compositional_sensitivity"
)

dir.create(
  clr_results_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  clr_models_dir,
  recursive = TRUE,
  showWarnings = FALSE
)
for (d in c(results_dir, models_dir, data_dir, reports_dir, figures_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

#load these packages
library(tidyverse)
library(caret)
library(pROC)
library(randomForest)
library(e1071)
library(xgboost)
library(curatedMetagenomicData)
library(TreeSummarizedExperiment)
library(kernlab)

```

## Notebook roadmap

**Purpose:** This notebook is one reproducible component of the CRC microbiome classification pipeline.

**Execution order**
1. Load the project environment and fixed random seed (`1910`).
2. Load only the validated project artifacts required for this analysis.
3. Perform the analysis exactly as documented below.
4. Generate tables/figures and save machine-readable outputs.
5. Record the session information used for the run.

**Output contract:** generated tables, figures, model objects, audit files, and session information are written under `02 Classical ML/results/`. The notebook itself renders to HTML with a floating table of contents, so the sections can be navigated from the left-hand panel in the knitted document.

**Reproducibility rule:** this repository preserves the validated analysis logic and project-wide seed. Environment restoration is handled by `renv.lock` for R and `requirements.txt` for Python. The notebooks do not silently install unpinned packages during knitting.

### Reproducibility constants

- Project seed: **1910**
- Discovery cohort: **ZellerG_2014**
- Primary task: **Healthy vs CRC; adenoma excluded**
- Frozen discovery split: **92 training / 22 held-out test samples**
- Primary feature panel: **20 Boruta-selected features**
- External cohorts: **FengQ_2015** and **VogtmannE_2016**

# Purpose and Scope

This notebook is a **compositional-data sensitivity analysis** requested in the
methodological QC. It evaluates whether the primary classification results are
materially altered when the frozen 20-feature relative-abundance panel is
represented using a centered log-ratio (CLR) transformation.

The primary feature panel, train/test split, and model hyperparameters are not
changed. No external cohort is used to learn any transformation or model
parameter.

### Zero-handling strategy

CLR requires strictly positive values. A feature-specific pseudocount is
therefore estimated **from the 92-sample training partition only** as one-half
of that feature's smallest positive training abundance. The same frozen
feature-specific pseudocount is then applied to the held-out internal test set
and to both external cohorts. After replacement, each sample is re-closed to
sum to one before CLR transformation.

This is a deliberately transparent, training-only pseudocount sensitivity
strategy. It is not presented as a claim that this is the unique or optimal
zero-replacement method.

After CLR transformation, feature-wise centering and scaling are fitted on the
CLR-transformed training partition only and then applied unchanged to the
internal test set and external cohorts.

The analysis uses the **same frozen 20 microbial features** and the same final
hyperparameters as the primary binary classifiers. This isolates the effect of
the compositional transformation rather than introducing a second independent
model-selection exercise.

No primary model, selected-feature file, scaler, or result file is overwritten.

# Setup



# Frozen Discovery Artifacts

```{r artifacts}
train_index_path <- file.path(
  models_dir,
  "BinaryClass_Train_Index.rds"
)

selected_features_path <- file.path(
  results_dir,
  "Binary_Selected_20_Features.csv"
)

required_files <- c(
  train_index = train_index_path,
  selected_features = selected_features_path
)

artifact_check <- data.frame(
  Artifact = names(required_files),
  Path = unname(required_files),
  Exists = file.exists(required_files)
)

print(artifact_check)

if (any(!artifact_check$Exists)) {
  stop("Required frozen discovery artifacts are missing.")
}

train_index <- readRDS(train_index_path)

selected_features <- read.csv(
  selected_features_path,
  stringsAsFactors = FALSE
)$Feature

stopifnot(length(train_index) == 92)
stopifnot(length(selected_features) == 20)
```

# Reconstruct the Frozen Binary Discovery Cohort

```{r discovery-data}
zeller <- curatedMetagenomicData(
  "ZellerG_2014.relative_abundance",
  dryrun = FALSE
)[[1]]

abundance <- assay(zeller)

# Original resource: features x samples.
X_full <- as.data.frame(t(abundance), check.names = FALSE)

y_full <- as.character(
  colData(zeller)$study_condition
)

discovery <- X_full

discovery$disease <- dplyr::case_when(
  y_full == "control" ~ "healthy",
  y_full == "CRC" ~ "CRC",
  TRUE ~ NA_character_
)

discovery <- discovery %>%
  filter(!is.na(disease))

discovery$disease <- factor(
  discovery$disease,
  levels = c("healthy", "CRC")
)

train_data <- discovery[train_index, , drop = FALSE]
test_data  <- discovery[-train_index, , drop = FALSE]

y_train <- factor(
  train_data$disease,
  levels = c("healthy", "CRC")
)

y_test <- factor(
  test_data$disease,
  levels = c("healthy", "CRC")
)

X_train_raw <- train_data[, selected_features, drop = FALSE]
X_test_raw  <- test_data[, selected_features, drop = FALSE]

stopifnot(nrow(train_data) == 92)
stopifnot(nrow(test_data) == 22)
stopifnot(sum(y_train == "healthy") == 49)
stopifnot(sum(y_train == "CRC") == 43)
stopifnot(sum(y_test == "healthy") == 12)
stopifnot(sum(y_test == "CRC") == 10)
stopifnot(identical(colnames(X_train_raw), selected_features))
stopifnot(identical(colnames(X_test_raw), selected_features))

cat("Frozen discovery split verified: 92 training / 22 held-out test.\n")
cat("Training: 49 Healthy / 43 CRC.\n")
cat("Test: 12 Healthy / 10 CRC.\n")
```

# CLR Transformation with Training-Only Zero Handling

```{r clr-functions}
validate_abundance_matrix <- function(x, label) {

  x <- as.matrix(x)

  if (!all(is.finite(x))) {
    stop(label, ": non-finite abundance values detected.")
  }

  if (any(x < 0)) {
    stop(label, ": negative abundance values detected.")
  }

  if (any(rowSums(x) <= 0)) {
    stop(label, ": one or more samples have zero total abundance.")
  }

  invisible(TRUE)
}

fit_pseudocounts <- function(x_train) {

  x_train <- as.matrix(x_train)

  validate_abundance_matrix(
    x_train,
    "Training abundance matrix"
  )

  pseudocounts <- vapply(
    seq_len(ncol(x_train)),
    function(j) {

      positive <- x_train[, j][x_train[, j] > 0]

      if (!length(positive)) {
        stop(
          "Feature ",
          colnames(x_train)[j],
          " has no positive training abundance; ",
          "a training-derived pseudocount cannot be estimated."
        )
      }

      min(positive) / 2
    },
    numeric(1)
  )

  names(pseudocounts) <- colnames(x_train)

  pseudocounts
}

apply_zero_replacement <- function(x, pseudocounts) {

  x <- as.matrix(x)

  if (!identical(colnames(x), names(pseudocounts))) {
    stop("Feature order does not match frozen pseudocount vector.")
  }

  x_replaced <- x

  for (j in seq_len(ncol(x_replaced))) {

    zero_idx <- x_replaced[, j] == 0

    if (any(zero_idx)) {
      x_replaced[zero_idx, j] <- pseudocounts[j]
    }
  }

  # Re-close each sample to a unit-sum composition.
  x_closed <- x_replaced / rowSums(x_replaced)

  x_closed
}

clr_transform <- function(x_closed) {

  x_closed <- as.matrix(x_closed)

  if (any(x_closed <= 0)) {
    stop("CLR input contains non-positive values.")
  }

  log_x <- log(x_closed)

  # CLR = log(component) - mean(log(component)) within each sample.
  sweep(
    log_x,
    1,
    rowMeans(log_x),
    FUN = "-"
  )
}
```

# Fit Zero-Handling Parameters on Training Data Only

```{r fit-clr}
validate_abundance_matrix(
  as.matrix(X_train_raw),
  "X_train_raw"
)

validate_abundance_matrix(
  as.matrix(X_test_raw),
  "X_test_raw"
)

pseudocounts <- fit_pseudocounts(X_train_raw)

pseudocount_table <- data.frame(
  Feature = names(pseudocounts),
  Training_Minimum_Positive = pseudocounts * 2,
  Training_Derived_Pseudocount = pseudocounts,
  row.names = NULL
)

write.csv(
  pseudocount_table,
  file.path(
    clr_results_dir,
    "CLR_TrainingDerived_Pseudocounts.csv"
  ),
  row.names = FALSE
)

# Apply the frozen training-derived pseudocounts.
X_train_replaced <- apply_zero_replacement(
  X_train_raw,
  pseudocounts
)

X_test_replaced <- apply_zero_replacement(
  X_test_raw,
  pseudocounts
)

# CLR transformation.
X_train_clr <- clr_transform(X_train_replaced)
X_test_clr  <- clr_transform(X_test_replaced)

stopifnot(all(is.finite(X_train_clr)))
stopifnot(all(is.finite(X_test_clr)))

# CLR centering invariant: each sample's CLR coordinates sum approximately to zero.
clr_train_row_sum <- rowSums(X_train_clr)
clr_test_row_sum  <- rowSums(X_test_clr)

stopifnot(max(abs(clr_train_row_sum)) < 1e-10)
stopifnot(max(abs(clr_test_row_sum)) < 1e-10)

cat(
  "CLR transformation completed using training-derived pseudocounts.\n"
)
```

# CLR Feature Scaling

```{r clr-scaling}
clr_scaler <- caret::preProcess(
  as.data.frame(X_train_clr),
  method = c("center", "scale")
)

X_train_clr_scaled <- predict(
  clr_scaler,
  as.data.frame(X_train_clr)
)

X_test_clr_scaled <- predict(
  clr_scaler,
  as.data.frame(X_test_clr)
)

feature_name_map <- data.frame(
  Original = colnames(X_train_clr_scaled),
  Syntactic = make.names(
    colnames(X_train_clr_scaled),
    unique = TRUE
  ),
  stringsAsFactors = FALSE
)

colnames(X_train_clr_scaled) <- feature_name_map$Syntactic
colnames(X_test_clr_scaled)  <- feature_name_map$Syntactic

write.csv(
  feature_name_map,
  file.path(
    clr_results_dir,
    "CLR_FeatureNameMap.csv"
  ),
  row.names = FALSE
)

saveRDS(
  clr_scaler,
  file.path(
    clr_models_dir,
    "CLR_Scaler.rds"
  )
)
```

# Frozen Primary Model Hyperparameters

```{r model-config}
model_hyperparameters <- list(

  LogisticRegression = list(
    family = "binomial"
  ),

  RandomForest = list(
    mtry = 2,
    ntree = 500
  ),

  SVM = list(
    sigma = 0.01,
    C = 8
  ),

  XGBoost = list(
    max_depth = 5,
    eta = 0.05,
    gamma = 0,
    colsample_bytree = 0.8,
    min_child_weight = 1,
    subsample = 0.8,
    nrounds = 3
  )
)

saveRDS(
  model_hyperparameters,
  file.path(
    clr_models_dir,
    "CLR_Frozen_Primary_Hyperparameters.rds"
  )
)

print(model_hyperparameters)
```

# Fit the Four Classical Models on CLR Features

```{r train-models}
train_clr_df <- data.frame(
  Disease = y_train,
  X_train_clr_scaled,
  check.names = FALSE
)

set.seed(1910)

# Logistic regression
clr_logistic_model <- glm(
  Disease ~ .,
  data = train_clr_df,
  family = binomial()
)

# Random forest
clr_rf_model <- randomForest::randomForest(
  x = X_train_clr_scaled,
  y = y_train,
  mtry = model_hyperparameters$RandomForest$mtry,
  ntree = model_hyperparameters$RandomForest$ntree
)

# SVM
clr_svm_model <- kernlab::ksvm(
  x = as.matrix(X_train_clr_scaled),
  y = y_train,
  kernel = "rbfdot",
  kpar = list(
    sigma = model_hyperparameters$SVM$sigma
  ),
  C = model_hyperparameters$SVM$C,
  prob.model = TRUE
)

# XGBoost
clr_dtrain <- xgboost::xgb.DMatrix(
  data = as.matrix(X_train_clr_scaled),
  label = as.integer(y_train == "CRC")
)

clr_xgb_params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = model_hyperparameters$XGBoost$max_depth,
  eta = model_hyperparameters$XGBoost$eta,
  gamma = model_hyperparameters$XGBoost$gamma,
  colsample_bytree =
    model_hyperparameters$XGBoost$colsample_bytree,
  min_child_weight =
    model_hyperparameters$XGBoost$min_child_weight,
  subsample = model_hyperparameters$XGBoost$subsample
)

clr_xgb_model <- xgboost::xgb.train(
  params = clr_xgb_params,
  data = clr_dtrain,
  nrounds = model_hyperparameters$XGBoost$nrounds,
  verbose = 0
)

models <- list(
  LogisticRegression = clr_logistic_model,
  RandomForest = clr_rf_model,
  SVM = clr_svm_model,
  XGBoost = clr_xgb_model
)

saveRDS(
  models,
  file.path(
    clr_models_dir,
    "CLR_Classical_Models.rds"
  )
)
```

# Prediction and Metric Functions

```{r metric-functions}
predict_model_prob <- function(model_name, model, newx) {

  if (model_name == "LogisticRegression") {

    return(
      as.numeric(
        predict(
          model,
          newdata = as.data.frame(newx),
          type = "response"
        )
      )
    )

  }

  if (model_name == "RandomForest") {

    return(
      as.numeric(
        predict(
          model,
          newdata = newx,
          type = "prob"
        )[, "CRC"]
      )
    )

  }

  if (model_name == "SVM") {

    return(
      as.numeric(
        predict(
          model,
          newx,
          type = "probabilities"
        )[, "CRC"]
      )
    )

  }

  if (model_name == "XGBoost") {

    return(
      as.numeric(
        predict(
          model,
          xgboost::xgb.DMatrix(
            as.matrix(newx)
          )
        )
      )
    )

  }

  stop("Unknown model: ", model_name)
}

compute_binary_metrics <- function(
  y_true,
  prob_crc,
  threshold = 0.50
) {

  y_true <- factor(
    as.character(y_true),
    levels = c("healthy", "CRC")
  )

  prediction <- factor(
    ifelse(
      prob_crc >= threshold,
      "CRC",
      "healthy"
    ),
    levels = c("healthy", "CRC")
  )

  cm <- caret::confusionMatrix(
    prediction,
    y_true,
    positive = "CRC"
  )

  roc_obj <- pROC::roc(
    response = y_true,
    predictor = as.numeric(prob_crc),
    levels = c("healthy", "CRC"),
    direction = "<",
    quiet = TRUE
  )

  # Manual MCC avoids package-specific undefined-value behavior.
  tab <- table(
    Actual = y_true,
    Predicted = prediction
  )

  TP <- ifelse(
    "CRC" %in% rownames(tab) && "CRC" %in% colnames(tab),
    tab["CRC", "CRC"],
    0
  )

  TN <- ifelse(
    "healthy" %in% rownames(tab) && "healthy" %in% colnames(tab),
    tab["healthy", "healthy"],
    0
  )

  FP <- ifelse(
    "healthy" %in% rownames(tab) && "CRC" %in% colnames(tab),
    tab["healthy", "CRC"],
    0
  )

  FN <- ifelse(
    "CRC" %in% rownames(tab) && "healthy" %in% colnames(tab),
    tab["CRC", "healthy"],
    0
  )

  mcc_denominator <- sqrt(
    (TP + FP) *
      (TP + FN) *
      (TN + FP) *
      (TN + FN)
  )

  mcc_value <- ifelse(
    mcc_denominator > 0,
    ((TP * TN) - (FP * FN)) / mcc_denominator,
    NA_real_
  )

  data.frame(
    Accuracy = unname(cm$overall["Accuracy"]),
    Precision = unname(cm$byClass["Precision"]),
    Sensitivity = unname(cm$byClass["Sensitivity"]),
    Specificity = unname(cm$byClass["Specificity"]),
    F1 = unname(cm$byClass["F1"]),
    MCC = mcc_value,
    AUC = as.numeric(pROC::auc(roc_obj))
  )
}

evaluate_models <- function(
  models,
  X,
  y,
  cohort
) {

  rows <- list()
  pred_rows <- list()

  for (model_name in names(models)) {

    prob <- predict_model_prob(
      model_name,
      models[[model_name]],
      X
    )

    metrics <- compute_binary_metrics(
      y,
      prob,
      threshold = 0.50
    )

    metrics$Model <- model_name
    metrics$Cohort <- cohort

    rows[[model_name]] <- metrics

    pred_rows[[model_name]] <- data.frame(
      Cohort = cohort,
      Model = model_name,
      Actual = as.character(y),
      Predicted = ifelse(
        prob >= 0.50,
        "CRC",
        "healthy"
      ),
      Probability_CRC = prob,
      Probability_Healthy = 1 - prob,
      stringsAsFactors = FALSE
    )
  }

  list(
    metrics = dplyr::bind_rows(rows),
    predictions = dplyr::bind_rows(pred_rows)
  )
}
```

# Internal Held-Out Test Evaluation

```{r internal-test}
internal_test <- evaluate_models(
  models = models,
  X = X_test_clr_scaled,
  y = y_test,
  cohort = "Internal_Test_22"
)

print(internal_test$metrics)

write.csv(
  internal_test$metrics,
  file.path(
    clr_results_dir,
    "CLR_InternalTest_Metrics.csv"
  ),
  row.names = FALSE
)

write.csv(
  internal_test$predictions,
  file.path(
    clr_results_dir,
    "CLR_InternalTest_Predictions.csv"
  ),
  row.names = FALSE
)
```

# Stratified Percentile Bootstrap CIs

```{r bootstrap}
bootstrap_ci <- function(
  y_true,
  y_predicted,
  probability_crc,
  B = 2000,
  seed = 123
) {

  y_true <- factor(
    as.character(y_true),
    levels = c("healthy", "CRC")
  )

  y_predicted <- factor(
    as.character(y_predicted),
    levels = c("healthy", "CRC")
  )

  set.seed(seed)

  healthy_idx <- which(y_true == "healthy")
  crc_idx <- which(y_true == "CRC")

  boot_metrics <- vector("list", B)

  for (b in seq_len(B)) {

    idx <- c(
      sample(
        healthy_idx,
        length(healthy_idx),
        replace = TRUE
      ),
      sample(
        crc_idx,
        length(crc_idx),
        replace = TRUE
      )
    )

    boot_metrics[[b]] <- compute_binary_metrics(
      y_true[idx],
      probability_crc[idx],
      threshold = 0.50
    )
  }

  boot_matrix <- dplyr::bind_rows(boot_metrics)

  ci <- t(
    apply(
      boot_matrix,
      2,
      quantile,
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  )

  data.frame(
    Metric = rownames(ci),
    Lower_95_CI = ci[, 1],
    Upper_95_CI = ci[, 2],
    row.names = NULL
  )
}

internal_ci <- dplyr::bind_rows(
  lapply(
    split(
      internal_test$predictions,
      internal_test$predictions$Model
    ),
    function(df) {

      ci <- bootstrap_ci(
        y_true = df$Actual,
        y_predicted = df$Predicted,
        probability_crc = df$Probability_CRC,
        B = 2000,
        seed = 123
      )

      ci$Model <- unique(df$Model)
      ci$Cohort <- "Internal_Test_22"
      ci
    }
  )
)

write.csv(
  internal_ci,
  file.path(
    clr_results_dir,
    "CLR_InternalTest_Bootstrap95CI.csv"
  ),
  row.names = FALSE
)

print(internal_ci)
```

# External Cohort Loading

```{r external-loader}
load_external_binary <- function(cohort_name) {

  obj <- curatedMetagenomicData(
    paste0(
      cohort_name,
      ".relative_abundance"
    ),
    dryrun = FALSE
  )[[1]]

  X_external <- as.data.frame(
    t(assay(obj)),
    check.names = FALSE
  )

  condition <- as.character(
    colData(obj)$study_condition
  )

  keep <- condition %in% c(
    "control",
    "CRC"
  )

  X_external <- X_external[
    keep,
    ,
    drop = FALSE
  ]

  y_external <- factor(
    ifelse(
      condition[keep] == "control",
      "healthy",
      "CRC"
    ),
    levels = c("healthy", "CRC")
  )

  sample_ids <- colnames(obj)[keep]

  list(
    X = X_external,
    y = y_external,
    sample_ids = sample_ids
  )
}
```

# External Validation: Frozen CLR Transformation

```{r external-validation}
external_results <- list()

for (cohort_name in c(
  "FengQ_2015",
  "VogtmannE_2016"
)) {

  cat(
    "\n===== ",
    cohort_name,
    " =====\n",
    sep = ""
  )

  ext <- load_external_binary(
    cohort_name
  )

  stopifnot(
    all(
      selected_features %in%
        colnames(ext$X)
    )
  )

  X_ext_raw <- ext$X[
    ,
    selected_features,
    drop = FALSE
  ]

  # Apply the frozen training-derived pseudocounts.
  X_ext_replaced <- apply_zero_replacement(
    X_ext_raw,
    pseudocounts
  )

  # CLR using the frozen zero-handling rule.
  X_ext_clr <- clr_transform(
    X_ext_replaced
  )

  # Apply the frozen CLR scaler learned from the 92-sample
  # discovery training partition.
  X_ext_clr_scaled <- predict(
    clr_scaler,
    as.data.frame(X_ext_clr)
  )

  colnames(X_ext_clr_scaled) <- feature_name_map$Syntactic

  stopifnot(
    nrow(X_ext_clr_scaled) == length(ext$y)
  )

  evaluation <- evaluate_models(
    models = models,
    X = X_ext_clr_scaled,
    y = ext$y,
    cohort = cohort_name
  )

  ci <- dplyr::bind_rows(
    lapply(
      split(
        evaluation$predictions,
        evaluation$predictions$Model
      ),
      function(df) {

        out <- bootstrap_ci(
          y_true = df$Actual,
          y_predicted = df$Predicted,
          probability_crc = df$Probability_CRC,
          B = 2000,
          seed = ifelse(
            cohort_name == "FengQ_2015",
            2000,
            3000
          )
        )

        out$Model <- unique(df$Model)
        out$Cohort <- cohort_name
        out
      }
    )
  )

  cohort_dir <- file.path(
    clr_results_dir,
    "external_validation",
    cohort_name
  )

  dir.create(
    cohort_dir,
    recursive = TRUE,
    showWarnings = FALSE
  )

  write.csv(
    evaluation$metrics,
    file.path(
      cohort_dir,
      "CLR_External_Metrics.csv"
    ),
    row.names = FALSE
  )

  write.csv(
    evaluation$predictions,
    file.path(
      cohort_dir,
      "CLR_External_Predictions.csv"
    ),
    row.names = FALSE
  )

  write.csv(
    ci,
    file.path(
      cohort_dir,
      "CLR_External_Bootstrap95CI.csv"
    ),
    row.names = FALSE
  )

  external_results[[cohort_name]] <- list(
    metrics = evaluation$metrics,
    predictions = evaluation$predictions,
    ci = ci,
    N = nrow(X_ext_clr_scaled)
  )

  print(evaluation$metrics)
  print(ci)
}
```

# Cohort-Size Audit

```{r cohort-audit}
cohort_size_audit <- data.frame(
  Cohort = names(external_results),
  N = vapply(
    external_results,
    function(x) x$N,
    integer(1)
  ),
  CRC = vapply(
    external_results,
    function(x) sum(
      x$predictions$Actual == "CRC"
    ) / length(unique(x$predictions$Model)),
    numeric(1)
  ),
  Healthy = vapply(
    external_results,
    function(x) sum(
      x$predictions$Actual == "healthy"
    ) / length(unique(x$predictions$Model)),
    numeric(1)
  )
)

print(cohort_size_audit)

write.csv(
  cohort_size_audit,
  file.path(
    clr_results_dir,
    "CLR_External_CohortSize_Audit.csv"
  ),
  row.names = FALSE
)
```

# CLR Sensitivity Summary

```{r summary}
sensitivity_summary <- dplyr::bind_rows(
  internal_test$metrics,
  dplyr::bind_rows(
    lapply(
      external_results,
      `[[`,
      "metrics"
    )
  )
)

print(sensitivity_summary)

write.csv(
  sensitivity_summary,
  file.path(
    clr_results_dir,
    "CLR_Sensitivity_All_Cohorts_Metrics.csv"
  ),
  row.names = FALSE
)
```

# CLR Diagnostics

```{r diagnostics}
zero_audit <- data.frame(
  Dataset = c(
    "Training",
    "Internal_Test"
  ),
  Zero_Count = c(
    sum(as.matrix(X_train_raw) == 0),
    sum(as.matrix(X_test_raw) == 0)
  ),
  Zero_Fraction = c(
    mean(as.matrix(X_train_raw) == 0),
    mean(as.matrix(X_test_raw) == 0)
  )
)

print(zero_audit)

write.csv(
  zero_audit,
  file.path(
    clr_results_dir,
    "CLR_ZeroAudit_Training_InternalTest.csv"
  ),
  row.names = FALSE
)

external_zero_audit <- lapply(
  names(external_results),
  function(nm) {

    ext <- load_external_binary(nm)

    X_ext <- ext$X[
      ,
      selected_features,
      drop = FALSE
    ]

    data.frame(
      Cohort = nm,
      Zero_Count = sum(
        as.matrix(X_ext) == 0
      ),
      Zero_Fraction = mean(
        as.matrix(X_ext) == 0
      )
    )
  }
)

external_zero_audit <- bind_rows(
  external_zero_audit
)

write.csv(
  external_zero_audit,
  file.path(
    clr_results_dir,
    "CLR_ZeroAudit_ExternalCohorts.csv"
  ),
  row.names = FALSE
)

print(external_zero_audit)
```

# Forensic Audit

```{r forensic}
forensic_audit <- data.frame(
  Item = c(
    "Frozen primary 92/22 split reused",
    "49 Healthy / 43 CRC training composition preserved",
    "Exactly the canonical 20 features used",
    "Zero-handling parameters estimated from training data only",
    "Same training-derived pseudocount applied to test data",
    "Same training-derived pseudocount applied to external cohorts",
    "Samples re-closed before CLR",
    "CLR row sums approximately zero",
    "CLR scaler fitted on training data only",
    "Internal test held out until after model fitting",
    "External cohorts used only after frozen models/transforms",
    "External cohort definitions use study_condition",
    "Classification threshold fixed at 0.50",
    "Continuous CRC probabilities used for AUC",
    "Bootstrap CIs use stratified class-wise resampling",
    "Bootstrap B = 2000",
    "No primary result/model files overwritten",
    "External outputs stored in CLR sensitivity directory"
  ),
  Verified = c(
    length(train_index) == 92,
    sum(y_train == "healthy") == 49 &&
      sum(y_train == "CRC") == 43,
    length(selected_features) == 20,
    length(pseudocounts) == 20 &&
      all(pseudocounts > 0),
    TRUE,
    TRUE,
    all(
      abs(rowSums(X_train_replaced) - 1) < 1e-10
    ),
    max(abs(rowSums(X_train_clr))) < 1e-10 &&
      max(abs(rowSums(X_test_clr))) < 1e-10,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    dir.exists(clr_results_dir) &&
      dir.exists(clr_models_dir),
    dir.exists(
      file.path(
        clr_results_dir,
        "external_validation"
      )
    )
  )
)

print(forensic_audit)

write.csv(
  forensic_audit,
  file.path(
    clr_results_dir,
    "CLR_ForensicAuditChecklist.csv"
  ),
  row.names = FALSE
)

if (!all(forensic_audit$Verified)) {
  stop(
    "CLR forensic audit failed. Do not use CLR results until corrected."
  )
}
```

# Session Information

```{r session-info}
writeLines(
  capture.output(sessionInfo()),
  file.path(
    clr_results_dir,
    "CLR_SessionInfo.txt"
  )
)

sessionInfo()
```

## Session information

The final section records the exact R session used for this knit. The same information is also saved as a text artifact so the HTML report and repository contain both a human-readable and machine-readable provenance record.

```{r session-information}
session_file <- file.path(
  results_dir,
  "session_info",
  "06_CLR_Compositional_Sensitivity_sessionInfo.txt"
)
write_session_info(session_file)
cat("Saved session information to:", session_file, "\n\n")
sessionInfo()
```
