---
title: "Nested Cross-Validation Sensitivity / Audit Analysis"
subtitle: "Healthy vs CRC Binary Classification — Fully Nested Preprocessing, Feature Selection and Tuning"
author: "Harshini Chintala"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    df_print: paged
    fig_caption: true
    self_contained: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "##",
  fig.width = 7.2,
  fig.height = 5.2,
  dpi = 300,
  fig.align = "center",
  out.width = "92%",
  results = "markup"
)

options(width = 90, scipen = 999)

source("../../R/project_setup.R")
set.seed(1910)

project_dir <- classical_dir
results_dir <- file.path(classical_dir, "results")
models_dir <- file.path(classical_dir, "models")
data_dir <- file.path(classical_dir, "data")
reports_dir <- file.path(classical_dir, "reports")
figures_dir <- file.path(classical_dir, "figures")

nested_root <- file.path(results_dir, "nested_cv")

for (d in c(
  results_dir,
  models_dir,
  data_dir,
  reports_dir,
  figures_dir,
  nested_root,
  file.path(nested_root, "metrics"),
  file.path(nested_root, "predictions"),
  file.path(nested_root, "feature_selection"),
  file.path(nested_root, "tuning"),
  file.path(models_dir, "nested_cv")
)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

library(caret)
library(dplyr)
library(pROC)
library(Boruta)
library(randomForest)
library(kernlab)
library(xgboost)
library(yardstick)
library(curatedMetagenomicData)
library(TreeSummarizedExperiment)

QUICK_TEST <- FALSE
BASE_SEED <- 1910

OUTER_K <- 10
OUTER_REPEATS <- 5
INNER_K <- 5

BORUTA_MAXRUNS <- 100

RF_NTREE <- 500
RF_MTRY_GRID <- 6

SVM_SIGMA <- c(0.01, 0.02)
SVM_C <- c(0.5, 1, 2, 4)

XGB_MAXROUNDS <- 1000
XGB_EARLYSTOP <- 20

XGB_GRID <- expand.grid(
  max_depth = c(2, 3, 5),
  eta = c(0.05, 0.1),
  gamma = 0,
  colsample_bytree = 0.7,
  min_child_weight = 1,
  subsample = c(0.8, 1.0)
)

```

## Notebook roadmap

**Purpose:** This notebook is one reproducible component of the CRC microbiome classification pipeline.

**Execution order**
1. Load the project environment and fixed random seed (`1910`).
2. Load only the validated project artifacts required for this analysis.
3. Perform the analysis exactly as documented below.
4. Generate tables/figures and save machine-readable outputs.
5. Record the session information used for the run.

**Output contract:** generated tables, figures, model objects, audit files, and session information are written under `02 Classical ML/results/`. The notebook itself renders to HTML with a floating table of contents, so the sections can be navigated from the left-hand panel in the knitted document.

**Reproducibility rule:** this repository preserves the validated analysis logic and project-wide seed. Environment restoration is handled by `renv.lock` for R and `requirements.txt` for Python. The notebooks do not silently install unpinned packages during knitting.

### Reproducibility constants

- Project seed: **1910**
- Discovery cohort: **ZellerG_2014**
- Primary task: **Healthy vs CRC; adenoma excluded**
- Frozen discovery split: **92 training / 22 held-out test samples**
- Primary feature panel: **20 Boruta-selected features**
- External cohorts: **FengQ_2015** and **VogtmannE_2016**

# Purpose

This notebook is a **methodological audit / sensitivity analysis** designed to
address the previously identified leakage risk in the primary classical ML
pipeline.

The key requirement is:

> Every data-dependent operation must be learned only from the data available
> inside the relevant training portion of the resampling procedure.

Therefore, in **every outer fold**:

1. NZV filtering is learned on outer-training data.
2. Boruta feature selection is learned on outer-training data.
3. Scaling parameters are learned on outer-training data.
4. Inside the outer-training data, the **inner CV** repeats steps 1–3 within
   each inner training fold before model tuning.
5. Hyperparameters are selected using only inner-validation predictions.
6. The final outer-fold model is then fitted on the **entire outer-training
   set**, using preprocessing/feature-selection rules learned from that entire
   outer-training set.
7. The resulting model is evaluated exactly once on the untouched outer
   validation fold.

The 22-sample primary held-out test set is **never used** in this notebook.

# Configuration



# Load Dataset and Reuse the Exact Primary 92/22 Split

```{r data}
zeller <- curatedMetagenomicData(
  "ZellerG_2014.relative_abundance", dryrun = FALSE
)[[1]]

abundance <- assay(zeller)
X_full <- as.data.frame(t(abundance))
y_full <- colData(zeller)$study_condition

crc_data <- X_full
crc_data$disease <- y_full

binary_data <- crc_data %>%
  mutate(disease = case_match(
    disease, "control" ~ "healthy", .default = disease
  )) %>%
  filter(disease %in% c("healthy", "CRC"))

binary_data$disease <- factor(
  binary_data$disease, levels = c("healthy", "CRC")
)

train_index_path <- file.path(models_dir, "BinaryClass_Train_Index.rds")
if (!file.exists(train_index_path)) {
  stop("Primary BinaryClass_Train_Index.rds not found.")
}

train_index <- readRDS(train_index_path)

train_data <- binary_data[train_index, ]
test_data <- binary_data[-train_index, ]

stopifnot(nrow(train_data) == 92, nrow(test_data) == 22)
stopifnot(sum(train_data$disease == "healthy") == 49)
stopifnot(sum(train_data$disease == "CRC") == 43)
stopifnot(sum(test_data$disease == "healthy") == 12)
stopifnot(sum(test_data$disease == "CRC") == 10)

X_pool <- train_data %>% select(-disease)
y_pool <- factor(train_data$disease, levels = c("healthy", "CRC"))

cat("Primary split reproduced exactly: 92 training / 22 held-out test.\n")
cat("The 22-sample test set is now locked and will not be used again.\n")
```

# Helper: Leakage-Safe Preprocessing

This function is the critical part of the audit.

It receives **training data and validation data for one resampling split**.
NZV filtering, Boruta, and scaling are learned from the training data only.

```{r preprocess-function}
preprocess_fold <- function(
  X_train, y_train, X_val,
  boruta_maxruns = 100,
  seed = 1910
) {

  set.seed(seed)

  # 1. NZV learned on training data only.
  nzv <- caret::nearZeroVar(X_train)

  if (length(nzv) > 0) {
    X_train_nzv <- X_train[, -nzv, drop = FALSE]
    X_val_nzv <- X_val[, -nzv, drop = FALSE]
  } else {
    X_train_nzv <- X_train
    X_val_nzv <- X_val
  }

  if (ncol(X_train_nzv) == 0) {
    stop("NZV filtering removed all predictors.")
  }

  # 2. Boruta learned on training data only.
  boruta_df <- data.frame(
    Disease = factor(y_train, levels = c("healthy", "CRC")),
    X_train_nzv,
    check.names = FALSE
  )

  boruta_fit <- tryCatch(
    Boruta::Boruta(
      Disease ~ .,
      data = boruta_df,
      doTrace = 0,
      maxRuns = boruta_maxruns
    ),
    error = function(e) NULL
  )

  fallback <- FALSE

  if (is.null(boruta_fit)) {
    selected <- colnames(X_train_nzv)
    fallback <- TRUE
  } else {
    boruta_fixed <- Boruta::TentativeRoughFix(boruta_fit)
    selected <- Boruta::getSelectedAttributes(
      boruta_fixed, withTentative = FALSE
    )

    if (length(selected) == 0) {
      selected <- colnames(X_train_nzv)
      fallback <- TRUE
    }
  }

  X_train_sel <- X_train_nzv[, selected, drop = FALSE]
  X_val_sel <- X_val_nzv[, selected, drop = FALSE]

  # 3. Scaling learned on training data only.
  scaler <- caret::preProcess(
    X_train_sel,
    method = c("center", "scale")
  )

  X_train_scaled <- predict(scaler, X_train_sel)
  X_val_scaled <- predict(scaler, X_val_sel)

  colnames(X_train_scaled) <- make.names(colnames(X_train_scaled))
  colnames(X_val_scaled) <- make.names(colnames(X_val_scaled))

  list(
    X_train = X_train_scaled,
    X_val = X_val_scaled,
    selected_features = selected,
    scaler = scaler,
    nzv_features = colnames(X_train)[nzv],
    boruta_fallback = fallback
  )
}
```

# Helper: Model Tuning Inside an Inner Fold

The function below receives the **raw outer-training subset** and performs
another complete resampling pipeline inside it.

For every inner split, preprocessing is refitted from that inner training
subset. This prevents inner validation observations from influencing feature
selection or scaling.

```{r inner-tuning}
tune_models_inner <- function(
  X_outer_train, y_outer_train,
  inner_k = 5,
  boruta_maxruns = 100,
  rf_ntree = 500,
  rf_mtry_grid = 6,
  svm_sigma = c(0.01, 0.02),
  svm_C = c(0.5, 1, 2, 4),
  xgb_grid = XGB_GRID,
  xgb_maxrounds = 1000,
  xgb_earlystop = 20,
  seed = 1910
) {

  set.seed(seed)

  inner_folds <- caret::createFolds(
    y_outer_train, k = inner_k,
    list = TRUE, returnTrain = FALSE
  )

  model_names <- c("BinaryLogistic", "RandomForest", "SVM", "XGBoost")
  score_store <- setNames(vector("list", length(model_names)), model_names)

  for (m in model_names) score_store[[m]] <- numeric()

  tuning_records <- list()

  for (j in seq_along(inner_folds)) {

    inner_val_idx <- inner_folds[[j]]
    inner_train_idx <- setdiff(
      seq_along(y_outer_train), inner_val_idx
    )

    prep <- preprocess_fold(
      X_outer_train[inner_train_idx, , drop = FALSE],
      y_outer_train[inner_train_idx],
      X_outer_train[inner_val_idx, , drop = FALSE],
      boruta_maxruns = boruta_maxruns,
      seed = seed + j
    )

    y_tr <- factor(
      y_outer_train[inner_train_idx],
      levels = c("healthy", "CRC")
    )
    y_va <- factor(
      y_outer_train[inner_val_idx],
      levels = c("healthy", "CRC")
    )

    df_tr <- data.frame(
      Disease = y_tr, prep$X_train, check.names = FALSE
    )

    # Logistic regression: no hyperparameters.
    lr <- tryCatch(
      glm(Disease ~ ., data = df_tr, family = binomial()),
      error = function(e) NULL
    )

    if (!is.null(lr)) {
      p <- as.numeric(
        predict(lr, newdata = prep$X_val, type = "response")
      )
      auc <- tryCatch(
        as.numeric(pROC::auc(
          pROC::roc(y_va, p, levels = c("healthy", "CRC"), quiet = TRUE)
        )),
        error = function(e) NA_real_
      )
      score_store$BinaryLogistic <- c(
        score_store$BinaryLogistic, auc
      )
    }

    # Random Forest tuning.
    rf_mtry <- unique(pmax(
      1,
      round(seq(
        1, ncol(prep$X_train),
        length.out = min(rf_mtry_grid, ncol(prep$X_train))
      ))
    ))

    rf_scores <- sapply(rf_mtry, function(mtry) {
      fit <- tryCatch(
        randomForest::randomForest(
          x = prep$X_train, y = y_tr,
          mtry = mtry, ntree = rf_ntree
        ),
        error = function(e) NULL
      )
      if (is.null(fit)) return(NA_real_)

      p <- predict(fit, prep$X_val, type = "prob")[, "CRC"]

      tryCatch(
        as.numeric(pROC::auc(
          pROC::roc(y_va, p, levels = c("healthy", "CRC"), quiet = TRUE)
        )),
        error = function(e) NA_real_
      )
    })

    best_rf <- rf_mtry[which.max(rf_scores)]
    score_store$RandomForest <- c(
      score_store$RandomForest,
      max(rf_scores, na.rm = TRUE)
    )

    tuning_records[[length(tuning_records) + 1]] <- data.frame(
      InnerFold = j,
      Model = "RandomForest",
      Best_mtry = best_rf
    )

    # SVM tuning.
    svm_grid <- expand.grid(
      sigma = svm_sigma, C = svm_C
    )

    svm_scores <- numeric(nrow(svm_grid))

    for (g in seq_len(nrow(svm_grid))) {
      fit <- tryCatch(
        kernlab::ksvm(
          x = as.matrix(prep$X_train),
          y = y_tr,
          kernel = "rbfdot",
          kpar = list(sigma = svm_grid$sigma[g]),
          C = svm_grid$C[g],
          prob.model = TRUE
        ),
        error = function(e) NULL
      )

      if (!is.null(fit)) {
        p <- tryCatch(
          as.numeric(
            predict(fit, prep$X_val, type = "probabilities")[, "CRC"]
          ),
          error = function(e) rep(NA_real_, length(y_va))
        )

        svm_scores[g] <- tryCatch(
          as.numeric(pROC::auc(
            pROC::roc(y_va, p,
                      levels = c("healthy", "CRC"), quiet = TRUE)
          )),
          error = function(e) NA_real_
        )
      }
    }

    best_svm <- svm_grid[which.max(svm_scores), , drop = FALSE]

    score_store$SVM <- c(
      score_store$SVM,
      max(svm_scores, na.rm = TRUE)
    )

    tuning_records[[length(tuning_records) + 1]] <- data.frame(
      InnerFold = j,
      Model = "SVM",
      Best_sigma = best_svm$sigma,
      Best_C = best_svm$C
    )

    # XGBoost tuning.
    dtr <- xgboost::xgb.DMatrix(
      as.matrix(prep$X_train),
      label = as.integer(y_tr == "CRC")
    )
    dva <- xgboost::xgb.DMatrix(
      as.matrix(prep$X_val),
      label = as.integer(y_va == "CRC")
    )

    xgb_scores <- numeric(nrow(xgb_grid))
    xgb_rounds <- numeric(nrow(xgb_grid))

    for (g in seq_len(nrow(xgb_grid))) {

      params <- list(
        objective = "binary:logistic",
        eval_metric = "auc",
        max_depth = xgb_grid$max_depth[g],
        eta = xgb_grid$eta[g],
        gamma = xgb_grid$gamma[g],
        colsample_bytree = xgb_grid$colsample_bytree[g],
        min_child_weight = xgb_grid$min_child_weight[g],
        subsample = xgb_grid$subsample[g]
      )

      fit <- tryCatch(
        xgboost::xgb.train(
          params = params,
          data = dtr,
          nrounds = xgb_maxrounds,
          watchlist = list(validation = dva),
          early_stopping_rounds = xgb_earlystop,
          verbose = 0
        ),
        error = function(e) NULL
      )

      if (!is.null(fit)) {
        p <- predict(fit, dva)

        xgb_scores[g] <- tryCatch(
          as.numeric(pROC::auc(
            pROC::roc(y_va, p,
                      levels = c("healthy", "CRC"), quiet = TRUE)
          )),
          error = function(e) NA_real_
        )

        # xgboost versions differ in whether `best_iteration` is returned
        # as a populated field. Never assign NULL into a preallocated vector.
        # If early stopping did not expose best_iteration, fall back to the
        # number of boosting iterations actually used.
        best_iter <- fit$best_iteration
        if (length(best_iter) == 0L || is.null(best_iter) ||
            !is.finite(as.numeric(best_iter))) {
          best_iter <- fit$niter
        }
        if (length(best_iter) == 0L || is.null(best_iter) ||
            !is.finite(as.numeric(best_iter))) {
          best_iter <- xgb_maxrounds
        }
        xgb_rounds[g] <- as.numeric(best_iter)
      }
    }

    if (all(is.na(xgb_scores))) {
      stop(
        "All XGBoost inner-validation AUC values were NA in inner fold ",
        j, ". Check xgboost installation/data before continuing."
      )
    }

    best_xgb_idx <- which.max(xgb_scores)
    best_xgb <- xgb_grid[best_xgb_idx, , drop = FALSE]

    score_store$XGBoost <- c(
      score_store$XGBoost,
      xgb_scores[best_xgb_idx]
    )

    tuning_records[[length(tuning_records) + 1]] <- data.frame(
      InnerFold = j,
      Model = "XGBoost",
      max_depth = best_xgb$max_depth,
      eta = best_xgb$eta,
      gamma = best_xgb$gamma,
      colsample_bytree = best_xgb$colsample_bytree,
      min_child_weight = best_xgb$min_child_weight,
      subsample = best_xgb$subsample,
      best_nrounds = xgb_rounds[best_xgb_idx]
    )
  }

  mean_scores <- sapply(score_store, function(x) {
    if (!length(x) || all(is.na(x))) NA_real_
    else mean(x, na.rm = TRUE)
  })

  if (all(is.na(mean_scores))) {
    stop("No model produced a valid inner-CV AUC in this outer fold.")
  }

  best_model <- names(which.max(mean_scores))

  list(
    best_model = best_model,
    mean_inner_auc = mean_scores,
    tuning_records = dplyr::bind_rows(tuning_records)
  )
}
```

# Outer Fold Evaluation

This is the complete nested architecture.

The outer validation fold is not touched by any preprocessing or tuning step
until the final prediction.

```{r outer-run}
run_outer_fold <- function(
  outer_train_idx, outer_val_idx,
  repeat_id, outer_fold_id
) {

  seed <- BASE_SEED + repeat_id * 1000 + outer_fold_id

  X_outer_train <- X_pool[outer_train_idx, , drop = FALSE]
  X_outer_val <- X_pool[outer_val_idx, , drop = FALSE]
  y_outer_train <- y_pool[outer_train_idx]
  y_outer_val <- y_pool[outer_val_idx]

  # ----------------------------------------------------------
  # INNER CV: all preprocessing is nested inside inner folds.
  # ----------------------------------------------------------
  inner_tuning <- tune_models_inner(
    X_outer_train,
    y_outer_train,
    inner_k = INNER_K,
    boruta_maxruns = BORUTA_MAXRUNS,
    rf_ntree = RF_NTREE,
    rf_mtry_grid = RF_MTRY_GRID,
    svm_sigma = SVM_SIGMA,
    svm_C = SVM_C,
    xgb_grid = XGB_GRID,
    xgb_maxrounds = XGB_MAXROUNDS,
    xgb_earlystop = XGB_EARLYSTOP,
    seed = seed
  )

  # ----------------------------------------------------------
  # FINAL OUTER-TRAIN PREPROCESSING.
  # This is fitted on ALL outer-training observations only.
  # ----------------------------------------------------------
  prep_outer <- preprocess_fold(
    X_outer_train,
    y_outer_train,
    X_outer_val,
    boruta_maxruns = BORUTA_MAXRUNS,
    seed = seed
  )

  y_tr <- factor(y_outer_train, levels = c("healthy", "CRC"))
  y_va <- factor(y_outer_val, levels = c("healthy", "CRC"))

  predictions <- list()

  # Model 1: Logistic regression.
  lr <- tryCatch(
    glm(
      Disease ~ .,
      data = data.frame(
        Disease = y_tr, prep_outer$X_train,
        check.names = FALSE
      ),
      family = binomial()
    ),
    error = function(e) NULL
  )

  if (!is.null(lr)) {
    p <- as.numeric(
      predict(lr, newdata = prep_outer$X_val, type = "response")
    )
    predictions$BinaryLogistic <- p
  } else {
    predictions$BinaryLogistic <- rep(NA_real_, length(y_va))
  }

  # Models are refitted on the complete outer-training set using
  # hyperparameters selected by inner CV.

  # Random Forest: select the best mtry observed during inner CV.
  # Use the modal mtry across inner folds, with deterministic tie-breaking.
  rf_records <- inner_tuning$tuning_records %>%
    filter(Model == "RandomForest")

  rf_mtry <- as.integer(
    names(sort(table(rf_records$Best_mtry), decreasing = TRUE))[1]
  )

  rf <- tryCatch(
    randomForest::randomForest(
      x = prep_outer$X_train,
      y = y_tr,
      mtry = rf_mtry,
      ntree = RF_NTREE
    ),
    error = function(e) NULL
  )

  predictions$RandomForest <- if (!is.null(rf)) {
    predict(rf, prep_outer$X_val, type = "prob")[, "CRC"]
  } else {
    rep(NA_real_, length(y_va))
  }

  # SVM: modal hyperparameter pair across inner folds.
  svm_records <- inner_tuning$tuning_records %>%
    filter(Model == "SVM")

  svm_counts <- svm_records %>%
    dplyr::count(Best_sigma, Best_C, sort = TRUE)

  if (nrow(svm_counts) == 0) {
    stop("No valid SVM tuning records were produced in this outer fold.")
  }

  best_svm <- svm_counts[1, , drop = FALSE]

  svm <- tryCatch(
    kernlab::ksvm(
      x = as.matrix(prep_outer$X_train),
      y = y_tr,
      kernel = "rbfdot",
      kpar = list(sigma = best_svm$Best_sigma),
      C = best_svm$Best_C,
      prob.model = TRUE
    ),
    error = function(e) NULL
  )

  predictions$SVM <- if (!is.null(svm)) {
    tryCatch(
      as.numeric(
        predict(
          svm, prep_outer$X_val,
          type = "probabilities"
        )[, "CRC"]
      ),
      error = function(e) rep(NA_real_, length(y_va))
    )
  } else {
    rep(NA_real_, length(y_va))
  }

  # XGBoost: modal parameter configuration, with median best iteration.
  xgb_records <- inner_tuning$tuning_records %>%
    filter(Model == "XGBoost")

  xgb_counts <- xgb_records %>%
    dplyr::count(
      max_depth, eta, gamma, colsample_bytree,
      min_child_weight, subsample,
      sort = TRUE
    )

  if (nrow(xgb_counts) == 0) {
    stop("No valid XGBoost tuning records were produced in this outer fold.")
  }

  best_xgb <- xgb_counts[1, , drop = FALSE]

  nrounds <- round(median(
    xgb_records$best_nrounds[
      xgb_records$max_depth == best_xgb$max_depth &
      xgb_records$eta == best_xgb$eta &
      xgb_records$gamma == best_xgb$gamma &
      xgb_records$colsample_bytree == best_xgb$colsample_bytree &
      xgb_records$min_child_weight == best_xgb$min_child_weight &
      xgb_records$subsample == best_xgb$subsample
    ],
    na.rm = TRUE
  ))

  if (!is.finite(nrounds) || nrounds < 1) nrounds <- 1

  params <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    max_depth = best_xgb$max_depth,
    eta = best_xgb$eta,
    gamma = best_xgb$gamma,
    colsample_bytree = best_xgb$colsample_bytree,
    min_child_weight = best_xgb$min_child_weight,
    subsample = best_xgb$subsample
  )

  xgb <- tryCatch(
    xgboost::xgb.train(
      params = params,
      data = xgboost::xgb.DMatrix(
        as.matrix(prep_outer$X_train),
        label = as.integer(y_tr == "CRC")
      ),
      nrounds = nrounds,
      verbose = 0
    ),
    error = function(e) NULL
  )

  predictions$XGBoost <- if (!is.null(xgb)) {
    predict(
      xgb,
      xgboost::xgb.DMatrix(as.matrix(prep_outer$X_val))
    )
  } else {
    rep(NA_real_, length(y_va))
  }

  metrics_for_model <- function(prob, model_name) {
    pred <- factor(
      ifelse(prob >= 0.50, "CRC", "healthy"),
      levels = c("healthy", "CRC")
    )

    cm <- tryCatch(
      caret::confusionMatrix(
        pred, y_va, positive = "CRC"
      ),
      error = function(e) NULL
    )

    auc <- tryCatch(
      as.numeric(pROC::auc(
        pROC::roc(
          y_va, prob,
          levels = c("healthy", "CRC"),
          quiet = TRUE
        )
      )),
      error = function(e) NA_real_
    )

    mcc <- tryCatch(
      as.numeric(
        yardstick::mcc(
          data.frame(truth = y_va, estimate = pred),
          truth = truth, estimate = estimate
        )$.estimate
      ),
      error = function(e) NA_real_
    )

    data.frame(
      Repeat = repeat_id,
      OuterFold = outer_fold_id,
      Model = model_name,
      N_OuterTrain = length(outer_train_idx),
      N_OuterVal = length(outer_val_idx),
      N_Selected_Features = length(prep_outer$selected_features),
      AUC = auc,
      Accuracy = if (!is.null(cm)) cm$overall["Accuracy"] else NA_real_,
      Sensitivity = if (!is.null(cm)) cm$byClass["Sensitivity"] else NA_real_,
      Specificity = if (!is.null(cm)) cm$byClass["Specificity"] else NA_real_,
      Precision = if (!is.null(cm)) cm$byClass["Precision"] else NA_real_,
      F1 = if (!is.null(cm)) cm$byClass["F1"] else NA_real_,
      MCC = mcc
    )
  }

  metric_rows <- list()
  prediction_rows <- list()

  for (current_model in names(predictions)) {
    prob <- predictions[[current_model]]

    metric_rows[[current_model]] <- metrics_for_model(prob, current_model)

    prediction_rows[[current_model]] <- data.frame(
      Repeat = repeat_id,
      OuterFold = outer_fold_id,
      Model = current_model,
      SampleID = rownames(X_outer_val),
      Actual = as.character(y_va),
      Probability_CRC = prob,
      Predicted = ifelse(prob >= 0.50, "CRC", "healthy")
    )
  }

  feature_row <- data.frame(
    Repeat = repeat_id,
    OuterFold = outer_fold_id,
    N_Selected_Features = length(prep_outer$selected_features),
    Selected_Features = paste(
      prep_outer$selected_features, collapse = "; "
    ),
    Boruta_Fallback = prep_outer$boruta_fallback
  )

  list(
    metrics = bind_rows(metric_rows),
    predictions = bind_rows(prediction_rows),
    feature_selection = feature_row,
    tuning = inner_tuning$tuning_records
  )
}
```

# Outer Resampling Plan

```{r outer-folds}
set.seed(BASE_SEED)

outer_results <- list()
counter <- 0

for (r in seq_len(OUTER_REPEATS)) {

  outer_folds <- caret::createFolds(
    y_pool, k = OUTER_K,
    list = TRUE, returnTrain = FALSE
  )

  for (f in seq_len(OUTER_K)) {

    counter <- counter + 1

    cat(
      "\n===== Outer evaluation ",
      counter, " / ",
      OUTER_K * OUTER_REPEATS,
      " | Repeat ", r,
      " Fold ", f, " =====\n",
      sep = ""
    )

    val_idx <- outer_folds[[f]]
    train_idx <- setdiff(seq_along(y_pool), val_idx)

    result <- run_outer_fold(
      train_idx, val_idx,
      repeat_id = r,
      outer_fold_id = f
    )

    outer_results[[counter]] <- result
  }
}

cat("\nCompleted ", length(outer_results), " outer evaluations.\n")
```

# Save Outer-Fold Results

```{r save-results}
all_metrics <- bind_rows(
  lapply(outer_results, `[[`, "metrics")
)

all_predictions <- bind_rows(
  lapply(outer_results, `[[`, "predictions")
)

all_features <- bind_rows(
  lapply(outer_results, `[[`, "feature_selection")
)

all_tuning <- bind_rows(
  lapply(outer_results, `[[`, "tuning")
)

write.csv(
  all_metrics,
  file.path(nested_root, "metrics", "NestedCV_OuterFold_Metrics.csv"),
  row.names = FALSE
)

write.csv(
  all_predictions,
  file.path(nested_root, "predictions", "NestedCV_OuterFold_Predictions.csv"),
  row.names = FALSE
)

write.csv(
  all_features,
  file.path(nested_root, "feature_selection", "NestedCV_FeatureSelection_ByOuterFold.csv"),
  row.names = FALSE
)

write.csv(
  all_tuning,
  file.path(nested_root, "tuning", "NestedCV_InnerTuning_Records.csv"),
  row.names = FALSE
)
```

# Nested-CV Summary

```{r summary}
nested_summary <- all_metrics %>%
  group_by(Model) %>%
  summarise(
    N_Outer_Evaluations = n(),
    Mean_AUC = mean(AUC, na.rm = TRUE),
    SD_AUC = sd(AUC, na.rm = TRUE),
    Mean_Accuracy = mean(Accuracy, na.rm = TRUE),
    SD_Accuracy = sd(Accuracy, na.rm = TRUE),
    Mean_Sensitivity = mean(Sensitivity, na.rm = TRUE),
    Mean_Specificity = mean(Specificity, na.rm = TRUE),
    Mean_Precision = mean(Precision, na.rm = TRUE),
    Mean_F1 = mean(F1, na.rm = TRUE),
    Mean_MCC = mean(MCC, na.rm = TRUE),
    .groups = "drop"
  )

print(nested_summary)

write.csv(
  nested_summary,
  file.path(nested_root, "metrics", "NestedCV_Summary.csv"),
  row.names = FALSE
)

feature_summary <- all_features %>%
  summarise(
    Mean_Selected_Features = mean(N_Selected_Features),
    SD_Selected_Features = sd(N_Selected_Features),
    Boruta_Fallback_Folds = sum(Boruta_Fallback)
  )

print(feature_summary)

write.csv(
  feature_summary,
  file.path(nested_root, "feature_selection", "NestedCV_FeatureSelection_Summary.csv"),
  row.names = FALSE
)
```

# Forensic Audit

```{r forensic-audit}
audit <- data.frame(
  Item = c(
    "Exact primary 92/22 split reused",
    "22-sample primary test set never used in fitting/tuning",
    "Outer validation observations excluded from outer preprocessing",
    "NZV fitted within each relevant training split",
    "Boruta fitted within each relevant training split",
    "Scaling fitted within each relevant training split",
    "Inner validation observations excluded from inner preprocessing",
    "Hyperparameter tuning uses inner validation only",
    "Outer evaluation performed after all model decisions",
    "Outer validation predictions saved",
    "Feature-selection membership saved by outer fold",
    "Inner tuning records saved",
    "No primary results/models directory root overwritten"
  ),
  Verified = c(
    nrow(train_data) == 92 && nrow(test_data) == 22,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    file.exists(file.path(
      nested_root, "predictions", "NestedCV_OuterFold_Predictions.csv"
    )),
    file.exists(file.path(
      nested_root, "feature_selection",
      "NestedCV_FeatureSelection_ByOuterFold.csv"
    )),
    file.exists(file.path(
      nested_root, "tuning", "NestedCV_InnerTuning_Records.csv"
    )),
    TRUE
  )
)

print(audit)

write.csv(
  audit,
  file.path(nested_root, "NestedCV_ForensicAuditChecklist.csv"),
  row.names = FALSE
)

if (!all(audit$Verified)) {
  stop("Nested-CV forensic audit failed.")
}
```



## Session information

The final section records the exact R session used for this knit. The same information is also saved as a text artifact so the HTML report and repository contain both a human-readable and machine-readable provenance record.

```{r session-information}
session_file <- file.path(
  results_dir,
  "session_info",
  "07_Nested_CV_Sensitivity_Audit_sessionInfo.txt"
)
write_session_info(session_file)
cat("Saved session information to:", session_file, "\n\n")
sessionInfo()
```
