---
title: "XGBoost Forensic Audit"
subtitle: "Independent audit of the original 48-configuration CV result"
author: "Harshini Chintala"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    df_print: paged
    fig_caption: true
    self_contained: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "##",
  fig.width = 7.2,
  fig.height = 5.2,
  dpi = 300,
  fig.align = "center",
  out.width = "92%",
  results = "markup"
)

options(width = 90, scipen = 999)
source("../../R/project_setup.R")
set.seed(1910)

# Classical notebook output locations.
project_dir <- classical_dir
results_dir <- file.path(classical_dir, "results")
models_dir <- file.path(classical_dir, "models")
data_dir <- file.path(classical_dir, "data")
reports_dir <- file.path(classical_dir, "reports")
figures_dir <- file.path(classical_dir, "figures")
for (d in c(results_dir, models_dir, data_dir, reports_dir, figures_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

# XGBoost forensic-audit output locations and execution controls.
audit_root <- file.path(
  results_dir,
  "XGBoost_Forensic_Audit"
)

dir.create(
  audit_root,
  recursive = TRUE,
  showWarnings = FALSE
)

checkpoint_path <- file.path(
  audit_root,
  "XGBoost_Manual_Rerun_Checkpoint.csv"
)

# TRUE = run/resume the complete independent 48-configuration audit.
# FALSE = only read an existing checkpoint and skip the expensive rerun.
RUN_FULL_MANUAL_RERUN <- TRUE

#load packages
library(dplyr)
library(pROC)
library(xgboost)
library(caret)
library(curatedMetagenomicData)
library(TreeSummarizedExperiment)

```

## Notebook roadmap

**Purpose:** This notebook is one reproducible component of the CRC microbiome classification pipeline.

**Execution order**
1. Load the project environment and fixed random seed (`1910`).
2. Load only the validated project artifacts required for this analysis.
3. Perform the analysis exactly as documented below.
4. Generate tables/figures and save machine-readable outputs.
5. Record the session information used for the run.

**Output contract:** generated tables, figures, model objects, audit files, and session information are written under `02 Classical ML/results/`. The notebook itself renders to HTML with a floating table of contents, so the sections can be navigated from the left-hand panel in the knitted document.

**Reproducibility rule:** this repository preserves the validated analysis logic and project-wide seed. Environment restoration is handled by `renv.lock` for R and `requirements.txt` for Python. The notebooks do not silently install unpinned packages during knitting.

### Reproducibility constants

- Project seed: **1910**
- Discovery cohort: **ZellerG_2014**
- Primary task: **Healthy vs CRC; adenoma excluded**
- Frozen discovery split: **92 training / 22 held-out test samples**
- Primary feature panel: **20 Boruta-selected features**
- External cohorts: **FengQ_2015** and **VogtmannE_2016**

# Purpose

This notebook audits the original binary XGBoost cross-validation result that
reported **CV AUC = 1.00 for all 48 hyperparameter configurations**.

The audit does not modify the primary model, feature set, train/test split,
or external-validation results.

It performs two levels of checking:

1. **Forensic integrity checks** on the saved fold definitions, labels,
   parameter grid, result table, sample IDs, and preprocessing artifacts.
2. **Independent manual out-of-fold reruns** using the saved hyperparameters
   and saved best boosting round for each configuration. For every
   configuration, each of the 50 held-out folds is predicted independently,
   fold-level ROC AUC is calculated with `pROC`, and both the mean fold AUC and
   pooled out-of-fold AUC are recorded.

This manual rerun is deliberately independent of `xgb.cv()`'s
`evaluation_log$test_auc_mean` extraction. The purpose is to determine whether
the originally reported perfect CV result is reproduced by an independent
fold-wise calculation.

# Configuration



# Load Saved Primary XGBoost Artifacts

```{r artifacts}
required_files <- c(
  train_index = file.path(
    models_dir,
    "BinaryClass_Train_Index.rds"
  ),
  cv_folds = file.path(
    models_dir,
    "BinaryClass_CV_Folds.rds"
  ),
  scaler = file.path(
    models_dir,
    "Scaler_Binary_XGB.rds"
  ),
  selected_features = file.path(
    results_dir,
    "XGBoostBinary_Selected_Features.csv"
  ),
  tuning_results = file.path(
    results_dir,
    "XGBoostBinary_CV_TuningResults.csv"
  ),
  final_params = file.path(
    models_dir,
    "XGBoostBinary_Model_Params.rds"
  )
)

artifact_status <- data.frame(
  Artifact = names(required_files),
  Path = unname(required_files),
  Exists = file.exists(required_files)
)

print(artifact_status)

if (any(!artifact_status$Exists)) {
  stop("One or more required XGBoost audit artifacts are missing.")
}

train_index <- readRDS(
  required_files["train_index"]
)

cv_folds <- readRDS(
  required_files["cv_folds"]
)

scaler <- readRDS(
  required_files["scaler"]
)

selected_features <- read.csv(
  required_files["selected_features"],
  stringsAsFactors = FALSE
)$Feature

saved_tuning <- read.csv(
  required_files["tuning_results"],
  stringsAsFactors = FALSE
)

saved_final_params <- readRDS(
  required_files["final_params"]
)

stopifnot(length(train_index) == 92)
stopifnot(length(cv_folds) == 50)
stopifnot(length(selected_features) == 20)
stopifnot(nrow(saved_tuning) == 48)

cat("Saved XGBoost tuning configurations:", nrow(saved_tuning), "\n")
cat("Saved CV folds:", length(cv_folds), "\n")
```

# Reconstruct the Exact Discovery Training Representation

```{r reconstruct-data}
zeller <- curatedMetagenomicData(
  "ZellerG_2014.relative_abundance",
  dryrun = FALSE
)[[1]]

abundance <- assay(zeller)

X_full <- as.data.frame(
  t(abundance),
  check.names = FALSE
)

y_full <- as.character(
  colData(zeller)$study_condition
)

discovery <- X_full

discovery$disease <- dplyr::case_when(
  y_full == "control" ~ "healthy",
  y_full == "CRC" ~ "CRC",
  TRUE ~ NA_character_
)

discovery <- discovery %>%
  filter(!is.na(disease))

discovery$disease <- factor(
  discovery$disease,
  levels = c("healthy", "CRC")
)

train_data <- discovery[train_index, , drop = FALSE]
test_data <- discovery[-train_index, , drop = FALSE]

y_train <- factor(
  train_data$disease,
  levels = c("healthy", "CRC")
)

y_test <- factor(
  test_data$disease,
  levels = c("healthy", "CRC")
)

X_train_raw <- train_data[
  ,
  setdiff(colnames(train_data), "disease"),
  drop = FALSE
]

# The saved Boruta panel is already a subset of the post-NZV feature space.
# Reconstruct the exact 20-feature training matrix and apply the saved scaler.
stopifnot(
  all(selected_features %in% colnames(X_train_raw))
)

X_train_selected <- X_train_raw[
  ,
  selected_features,
  drop = FALSE
]

X_train_scaled <- predict(
  scaler,
  X_train_selected
)

colnames(X_train_scaled) <- make.names(
  colnames(X_train_scaled),
  unique = TRUE
)

stopifnot(
  nrow(X_train_scaled) == 92,
  ncol(X_train_scaled) == 20,
  sum(y_train == "healthy") == 49,
  sum(y_train == "CRC") == 43
)

train_matrix <- as.matrix(
  X_train_scaled
)

cat("Training matrix:", nrow(train_matrix), "x", ncol(train_matrix), "\n")
cat("Training class balance: Healthy =", sum(y_train == "healthy"),
    "; CRC =", sum(y_train == "CRC"), "\n")
```

# Audit 1: Sample-ID and Label Integrity

```{r sample-id-audit}
sample_ids <- rownames(train_data)

sample_id_audit <- data.frame(
  N_samples = length(sample_ids),
  N_unique_sample_ids = length(unique(sample_ids)),
  Any_duplicate_sample_ids = anyDuplicated(sample_ids) > 0,
  Labels_length_matches = length(y_train) == length(sample_ids),
  Row_order_reproducible = identical(
    rownames(X_train_scaled),
    sample_ids
  )
)

print(sample_id_audit)

stopifnot(
  sample_id_audit$N_samples == 92,
  sample_id_audit$N_unique_sample_ids == 92,
  !sample_id_audit$Any_duplicate_sample_ids,
  sample_id_audit$Labels_length_matches
)
```

# Audit 2: Cross-Validation Fold Structure

```{r fold-audit}
fold_names <- names(cv_folds)

fold_structure <- data.frame(
  Fold_Name = fold_names,
  Training_Size = vapply(
    cv_folds,
    length,
    integer(1)
  )
)

fold_structure$Validation_Size <- 92 -
  fold_structure$Training_Size

print(fold_structure)

stopifnot(
  length(cv_folds) == 50,
  all(fold_structure$Training_Size > 0),
  all(fold_structure$Validation_Size > 0)
)
```

```{r repeat-audit}
extract_repeat <- function(x) {
  as.integer(
    sub(
      "^Fold[0-9]+\\.Rep([0-9]+)$",
      "\\1",
      x
    )
  )
}

extract_fold <- function(x) {
  as.integer(
    sub(
      "^Fold([0-9]+)\\.Rep[0-9]+$",
      "\\1",
      x
    )
  )
}

repeat_ids <- extract_repeat(fold_names)
fold_ids <- extract_fold(fold_names)

repeat_audit <- do.call(
  rbind,
  lapply(
    sort(unique(repeat_ids)),
    function(r) {

      idx <- which(repeat_ids == r)

      validation_indices <- lapply(
        idx,
        function(i) {
          setdiff(
            seq_len(92),
            cv_folds[[i]]
          )
        }
      )

      counts <- tabulate(
        unlist(validation_indices),
        nbins = 92
      )

      data.frame(
        Repeat = r,
        Number_Folds = length(idx),
        Every_Sample_Validated_Once = all(counts == 1),
        Minimum_Class_Count = min(
          vapply(
            validation_indices,
            function(v) {
              min(
                table(
                  y_train[v]
                )
              )
            },
            numeric(1)
          )
        ),
        Maximum_Validation_Size = max(
          vapply(
            validation_indices,
            length,
            integer(1)
          )
        )
      )
    }
  )
)

print(repeat_audit)

stopifnot(
  all(repeat_audit$Number_Folds == 10),
  all(repeat_audit$Every_Sample_Validated_Once),
  all(repeat_audit$Minimum_Class_Count >= 1)
)
```

# Audit 3: Saved Parameter Grid

```{r grid-audit}
expected_grid <- expand.grid(
  max_depth        = c(2, 3, 5),
  eta              = c(0.05, 0.1),
  gamma            = c(0, 0.5),
  colsample_bytree = c(0.5, 0.8),
  min_child_weight = 1,
  subsample        = c(0.8, 1.0)
)

parameter_columns <- c(
  "max_depth",
  "eta",
  "gamma",
  "colsample_bytree",
  "min_child_weight",
  "subsample"
)

saved_key <- do.call(
  paste,
  c(
    saved_tuning[, parameter_columns],
    sep = "|"
  )
)

expected_key <- do.call(
  paste,
  c(
    expected_grid[, parameter_columns],
    sep = "|"
  )
)

grid_audit <- data.frame(
  Expected_Configurations = nrow(expected_grid),
  Saved_Configurations = nrow(saved_tuning),
  Saved_Unique_Configurations =
    length(unique(saved_key)),
  Grid_Set_Matches =
    setequal(saved_key, expected_key),
  All_Saved_AUCs_Equal_1 =
    all(saved_tuning$best_cv_auc == 1),
  Minimum_Saved_AUC =
    min(saved_tuning$best_cv_auc),
  Maximum_Saved_AUC =
    max(saved_tuning$best_cv_auc)
)

print(grid_audit)

stopifnot(
  grid_audit$Expected_Configurations == 48,
  grid_audit$Saved_Configurations == 48,
  grid_audit$Saved_Unique_Configurations == 48,
  grid_audit$Grid_Set_Matches,
  grid_audit$Minimum_Saved_AUC == 1,
  grid_audit$Maximum_Saved_AUC == 1
)
```

# Audit 4: Check the Saved Final Hyperparameters

```{r final-parameter-audit}
print(saved_final_params)

# The primary XGBoost notebook saves:
#   list(params = final_params, nrounds = best_nrounds)
# so the model parameters live inside saved_final_params$params.
saved_model_params <- saved_final_params$params
saved_nrounds <- saved_final_params$nrounds

expected_param_names <- c(
  "max_depth",
  "eta",
  "gamma",
  "colsample_bytree",
  "min_child_weight",
  "subsample"
)

final_parameter_audit <- data.frame(
  Parameter_Object_Loaded = !is.null(saved_final_params),
  Has_Model_Parameters = !is.null(saved_model_params) &&
    all(expected_param_names %in% names(saved_model_params)),
  Has_Best_Nrounds = !is.null(saved_nrounds) &&
    length(saved_nrounds) == 1 &&
    is.finite(saved_nrounds) &&
    saved_nrounds >= 1
)

print(final_parameter_audit)

if (!all(final_parameter_audit)) {
  stop(
    "Saved final XGBoost parameter object does not contain the expected ",
    "nested 'params' and 'nrounds' fields."
  )
}

# Make the extracted values explicit for the rest of the audit.
final_parameter_vector <- c(
  max_depth = as.numeric(saved_model_params$max_depth),
  eta = as.numeric(saved_model_params$eta),
  gamma = as.numeric(saved_model_params$gamma),
  colsample_bytree = as.numeric(saved_model_params$colsample_bytree),
  min_child_weight = as.numeric(saved_model_params$min_child_weight),
  subsample = as.numeric(saved_model_params$subsample),
  nrounds = as.numeric(saved_nrounds)
)

print(final_parameter_vector)
```

# Audit 5: Check for Training/Validation Overlap

```{r overlap-audit}
overlap_audit <- lapply(
  seq_along(cv_folds),
  function(i) {

    tr <- cv_folds[[i]]
    va <- setdiff(
      seq_len(92),
      tr
    )

    list(
      Fold = fold_names[i],
      Any_Overlap = length(intersect(tr, va)) > 0,
      Union_Covers_All = length(
        union(tr, va)
      ) == 92
    )
  }
)

overlap_table <- data.frame(
  Fold = vapply(
    overlap_audit,
    `[[`,
    character(1),
    "Fold"
  ),
  Any_Overlap = vapply(
    overlap_audit,
    `[[`,
    logical(1),
    "Any_Overlap"
  ),
  Union_Covers_All = vapply(
    overlap_audit,
    `[[`,
    logical(1),
    "Union_Covers_All"
  )
)

print(overlap_table)

stopifnot(
  !any(overlap_table$Any_Overlap),
  all(overlap_table$Union_Covers_All)
)
```

# Independent Manual Out-of-Fold Rerun

This is the core forensic test.

For each saved configuration, the reported `best_nrounds` is used. Each of the
50 held-out folds is fitted independently with `xgb.train()`. The held-out
probabilities are passed to `pROC` to calculate ROC AUC outside `xgb.cv()`.

The process is checkpointed after every configuration so that a long audit can
resume rather than losing all completed work if the session is interrupted.

```{r manual-rerun}
manual_one_configuration <- function(
  config_row,
  X,
  y,
  cv_folds,
  seed_base = 1910
) {

  params <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    max_depth = as.integer(config_row$max_depth),
    eta = as.numeric(config_row$eta),
    gamma = as.numeric(config_row$gamma),
    colsample_bytree = as.numeric(
      config_row$colsample_bytree
    ),
    min_child_weight = as.numeric(
      config_row$min_child_weight
    ),
    subsample = as.numeric(
      config_row$subsample
    )
  )

  nrounds <- as.integer(
    config_row$best_nrounds
  )

  fold_aucs <- numeric(
    length(cv_folds)
  )

  # createMultiFolds() produces training indices. Their complements are
  # validation folds. With 5 repeats, each sample is therefore held out
  # exactly 5 times, not once overall. Store all repeated held-out
  # probabilities and average them per sample for a pooled repeated-CV AUC.
  pooled_probability_sum <- numeric(
    nrow(X)
  )

  pooled_seen <- integer(
    nrow(X)
  )

  for (i in seq_along(cv_folds)) {

    validation_idx <- setdiff(
      seq_len(nrow(X)),
      cv_folds[[i]]
    )

    training_idx <- cv_folds[[i]]

    if (length(unique(y[validation_idx])) != 2) {
      stop(
        "Validation fold ", i,
        " does not contain both classes."
      )
    }

    dtrain <- xgboost::xgb.DMatrix(
      data = X[training_idx, , drop = FALSE],
      label = as.integer(
        y[training_idx] == "CRC"
      )
    )

    dvalidation <- xgboost::xgb.DMatrix(
      data = X[validation_idx, , drop = FALSE],
      label = as.integer(
        y[validation_idx] == "CRC"
      )
    )

    set.seed(
      seed_base +
        i +
        as.integer(config_row$max_depth) * 100 +
        round(config_row$eta * 1000)
    )

    fit <- xgboost::xgb.train(
      params = params,
      data = dtrain,
      nrounds = nrounds,
      verbose = 0
    )

    prob <- as.numeric(
      predict(
        fit,
        dvalidation
      )
    )

    fold_aucs[i] <- as.numeric(
      pROC::auc(
        pROC::roc(
          response = y[validation_idx],
          predictor = prob,
          levels = c("healthy", "CRC"),
          direction = "<",
          quiet = TRUE
        )
      )
    )

    pooled_probability_sum[validation_idx] <-
      pooled_probability_sum[validation_idx] + prob

    pooled_seen[validation_idx] <-
      pooled_seen[validation_idx] + 1L
  }

  # Five repeats mean each sample has five independent held-out
  # predictions. Average those five probabilities before calculating
  # the pooled repeated-CV AUC.
  expected_repeats <- 5L

  if (!all(pooled_seen == expected_repeats)) {
    stop(
      "Repeated out-of-fold predictions do not cover each sample ",
      expected_repeats,
      " times as expected. Observed counts: ",
      paste(
        sort(unique(pooled_seen)),
        collapse = ", "
      )
    )
  }

  pooled_probability <-
    pooled_probability_sum / pooled_seen

  pooled_auc <- as.numeric(
    pROC::auc(
      pROC::roc(
        response = y,
        predictor = pooled_probability,
        levels = c("healthy", "CRC"),
        direction = "<",
        quiet = TRUE
      )
    )
  )

  data.frame(
    max_depth = config_row$max_depth,
    eta = config_row$eta,
    gamma = config_row$gamma,
    colsample_bytree = config_row$colsample_bytree,
    min_child_weight = config_row$min_child_weight,
    subsample = config_row$subsample,
    saved_best_nrounds = config_row$best_nrounds,
    saved_cv_auc = config_row$best_cv_auc,
    manual_mean_fold_auc = mean(fold_aucs),
    manual_sd_fold_auc = sd(fold_aucs),
    manual_min_fold_auc = min(fold_aucs),
    manual_max_fold_auc = max(fold_aucs),
    repeated_oof_predictions_per_sample = expected_repeats,
    manual_pooled_oof_auc = pooled_auc,
    all_fold_auc_equal_1 = all(
      abs(fold_aucs - 1) < 1e-12
    )
  )
}

if (RUN_FULL_MANUAL_RERUN) {

  if (file.exists(checkpoint_path)) {

    checkpoint <- read.csv(
      checkpoint_path,
      stringsAsFactors = FALSE
    )

  } else {

    checkpoint <- data.frame()

  }

  rerun_results <- list()

  for (i in seq_len(nrow(saved_tuning))) {

    cfg <- saved_tuning[i, ]

    # Build a single scalar key from the six tuning parameters.
    # Passing a data.frame directly to paste() can produce a vector,
    # which then makes `if (already_done)` fail with "condition has
    # length > 1". Explicit scalar extraction avoids that ambiguity.
    make_config_key <- function(row) {
      paste(
        as.character(row$max_depth),
        as.character(row$eta),
        as.character(row$gamma),
        as.character(row$colsample_bytree),
        as.character(row$min_child_weight),
        as.character(row$subsample),
        sep = "|"
      )
    }

    key <- make_config_key(cfg)

    already_done <- FALSE

    if (nrow(checkpoint) > 0) {

      checkpoint_keys <- vapply(
        seq_len(nrow(checkpoint)),
        function(j) {
          make_config_key(
            checkpoint[j, , drop = FALSE]
          )
        },
        character(1)
      )

      already_done <- isTRUE(
        key %in% checkpoint_keys
      )
    }

    if (already_done) {

      cat(
        "Skipping completed configuration ",
        i,
        "/",
        nrow(saved_tuning),
        "\n",
        sep = ""
      )

      next
    }

    cat(
      "\n===== Manual audit configuration ",
      i,
      " / ",
      nrow(saved_tuning),
      " =====\n",
      sep = ""
    )

    result <- manual_one_configuration(
      config_row = cfg,
      X = train_matrix,
      y = y_train,
      cv_folds = cv_folds,
      seed_base = 1910
    )

    checkpoint <- dplyr::bind_rows(
      checkpoint,
      result
    )

    write.csv(
      checkpoint,
      checkpoint_path,
      row.names = FALSE
    )

    cat(
      "Saved checkpoint after configuration ",
      i,
      ".\n",
      sep = ""
    )
  }

  manual_results <- checkpoint

} else {

  manual_results <- if (
    file.exists(checkpoint_path)
  ) {
    read.csv(
      checkpoint_path,
      stringsAsFactors = FALSE
    )
  } else {
    data.frame()
  }

}

if (RUN_FULL_MANUAL_RERUN && nrow(manual_results) != 48) {
  stop(
    "Full manual rerun did not produce all 48 configurations. ",
    "The checkpoint can be resumed."
  )
}

print(manual_results)
```

# Compare Original and Independent Manual AUC

```{r comparison}
if (nrow(manual_results) > 0) {

  comparison <- saved_tuning %>%
    left_join(
      manual_results,
      by = parameter_columns
    ) %>%
    mutate(
      Mean_Fold_AUC_Difference =
        manual_mean_fold_auc - best_cv_auc,
      Pooled_OOF_AUC_Difference =
        manual_pooled_oof_auc - best_cv_auc,
      Manual_Reproduces_Perfect_Mean =
        abs(
          manual_mean_fold_auc - 1
        ) < 1e-12,
      Manual_Reproduces_Perfect_Pooled =
        abs(
          manual_pooled_oof_auc - 1
        ) < 1e-12
    )

  write.csv(
    comparison,
    file.path(
      audit_root,
      "XGBoost_Original_vs_Manual_AUC_Comparison.csv"
    ),
    row.names = FALSE
  )

  print(
    comparison[
      ,
      c(
        parameter_columns,
        "saved_best_nrounds",
        "saved_cv_auc",
        "manual_mean_fold_auc",
        "manual_pooled_oof_auc",
        "Mean_Fold_AUC_Difference",
        "Pooled_OOF_AUC_Difference"
      )
    ]
  )
}
```

# Forensic Interpretation

```{r interpretation}
if (nrow(manual_results) == 48) {

  forensic_summary <- data.frame(
    Saved_Configurations = 48,
    Saved_All_AUC_Exactly_1 =
      all(saved_tuning$best_cv_auc == 1),
    Manual_Mean_AUC_All_Exactly_1 =
      all(
        abs(
          manual_results$manual_mean_fold_auc - 1
        ) < 1e-12
      ),
    Manual_Pooled_OOF_AUC_All_Exactly_1 =
      all(
        abs(
          manual_results$manual_pooled_oof_auc - 1
        ) < 1e-12
      ),
    Manual_Min_Mean_Fold_AUC =
      min(
        manual_results$manual_mean_fold_auc
      ),
    Manual_Max_Mean_Fold_AUC =
      max(
        manual_results$manual_mean_fold_auc
      ),
    Manual_Min_Pooled_OOF_AUC =
      min(
        manual_results$manual_pooled_oof_auc
      ),
    Manual_Max_Pooled_OOF_AUC =
      max(
        manual_results$manual_pooled_oof_auc
      )
  )

  print(forensic_summary)

  write.csv(
    forensic_summary,
    file.path(
      audit_root,
      "XGBoost_Forensic_Summary.csv"
    ),
    row.names = FALSE
  )
}
```

# Final Audit Checklist

```{r final-audit}
final_audit <- data.frame(
  Item = c(
    "Exactly 92 discovery-training samples reconstructed",
    "Training labels are 49 Healthy / 43 CRC",
    "Exactly 20 saved XGBoost features reconstructed",
    "No duplicated training sample IDs",
    "50 saved CV folds present",
    "10 folds in each of 5 repeats",
    "Each sample held out exactly once per repeat",
    "Every validation fold contains both classes",
    "No training/validation index overlap",
    "Saved parameter grid contains exactly 48 unique configurations",
    "Every saved original best CV AUC equals 1",
    "Saved final parameter object loaded",
    "Independent manual rerun completed for all 48 configurations",
    "Manual fold AUCs are finite and within [0, 1]"
  ),
  Verified = c(
    nrow(X_train_scaled) == 92,
    sum(y_train == "healthy") == 49 &&
      sum(y_train == "CRC") == 43,
    ncol(X_train_scaled) == 20,
    length(unique(sample_ids)) == 92,
    length(cv_folds) == 50,
    all(repeat_audit$Number_Folds == 10),
    all(repeat_audit$Every_Sample_Validated_Once),
    all(repeat_audit$Minimum_Class_Count >= 1),
    !any(overlap_table$Any_Overlap),
    grid_audit$Expected_Configurations == 48 &&
      grid_audit$Saved_Unique_Configurations == 48 &&
      grid_audit$Grid_Set_Matches,
    all(saved_tuning$best_cv_auc == 1),
    all(final_parameter_audit),
    nrow(manual_results) == 48,
    all(
      is.finite(
        manual_results$manual_mean_fold_auc
      )
    ) &&
      all(
        manual_results$manual_mean_fold_auc >= 0 &
          manual_results$manual_mean_fold_auc <= 1
      )
  )
)

print(final_audit)

write.csv(
  final_audit,
  file.path(
    audit_root,
    "XGBoost_Forensic_Audit_Checklist.csv"
  ),
  row.names = FALSE
)

if (!all(final_audit$Verified)) {
  stop(
    "Forensic audit checklist failed. Do not treat the original XGBoost CV ",
    "result as validated until the failed checks are resolved."
  )
}
```



## Session information

The final section records the exact R session used for this knit. The same information is also saved as a text artifact so the HTML report and repository contain both a human-readable and machine-readable provenance record.

```{r session-information}
session_file <- file.path(
  results_dir,
  "session_info",
  "08_XGBoost_Forensic_Audit_sessionInfo.txt"
)
write_session_info(session_file)
cat("Saved session information to:", session_file, "\n\n")
sessionInfo()
```
