---
title: "creating_quantum"
author: "Harshini"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    df_print: paged
    fig_caption: true
    self_contained: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "##",
  fig.width = 7.2,
  fig.height = 5.2,
  dpi = 300,
  fig.align = "center",
  out.width = "92%",
  results = "markup"
)

options(width = 90, scipen = 999)
source("../../R/project_setup.R")
set.seed(1910)

# Quantum notebook output locations. Existing validated artifacts are never overwritten
# unless the analysis code explicitly writes to those same documented paths.
results_dir <- file.path(quantum_dir, "results")
models_dir <- file.path(quantum_dir, "models")
data_dir <- file.path(quantum_dir, "data")
figures_dir <- file.path(quantum_dir, "figures")
for (d in c(results_dir, models_dir, data_dir, figures_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

```

## Notebook roadmap

**Purpose:** This notebook is one reproducible component of the CRC microbiome classification pipeline.

**Execution order**
1. Load the project environment and fixed random seed (`1910`).
2. Load only the validated project artifacts required for this analysis.
3. Perform the analysis exactly as documented below.
4. Generate tables/figures and save machine-readable outputs.
5. Record the session information used for the run.

**Output contract:** generated tables, figures, model objects, audit files, and session information are written under `03 Quantum_ML/05 QSVC/results/`. The notebook itself renders to HTML with a floating table of contents, so the sections can be navigated from the left-hand panel in the knitted document.

**Reproducibility rule:** this repository preserves the validated analysis logic and project-wide seed. Environment restoration is handled by `renv.lock` for R and `requirements.txt` for Python. The notebooks do not silently install unpinned packages during knitting.

### Reproducibility constants

- Project seed: **1910**
- Discovery cohort: **ZellerG_2014**
- Primary task: **Healthy vs CRC; adenoma excluded**
- Frozen discovery split: **92 training / 22 held-out test samples**
- Primary feature panel: **20 Boruta-selected features**
- External cohorts: **FengQ_2015** and **VogtmannE_2016**

```{r}
# ==========================================================
# QSVC DATA EXPORT
# Run this in your R environment
# ==========================================================

# ==========================================================
# 1. LOAD REQUIRED PACKAGES
# ==========================================================

library(curatedMetagenomicData)
library(TreeSummarizedExperiment)
library(SummarizedExperiment)
library(caret)
library(Boruta)
library(dplyr)
# ===========================================# ==========================================================
# PROJECT PATHS
# ==========================================================

# Locate the repository root from the directory used for knitting.
# This avoids hard-coded Windows paths and does not depend on
# project_setup.R's internal path variables.

find_project_root <- function(start_dir = getwd()) {

  current <- normalizePath(
    start_dir,
    winslash = "/",
    mustWork = TRUE
  )

  for (i in 0:10) {

    has_classical <- dir.exists(
      file.path(current, "02 Classical ML")
    )

    has_quantum <- dir.exists(
      file.path(current, "03 Quantum_ML")
    )

    if (has_classical && has_quantum) {
      return(current)
    }

    parent <- dirname(current)

    if (identical(parent, current)) {
      break
    }

    current <- parent
  }

  stop(
    paste0(
      "ERROR: Could not locate CRC project root from knitting directory.\n",
      "Starting directory: ", normalizePath(start_dir, winslash = "/", mustWork = TRUE), "\n",
      "Expected an ancestor containing both '02 Classical ML' and '03 Quantum_ML'."
    )
  )
}

project_dir <- find_project_root()

# Classical ML folder
classical_dir <- file.path(
  project_dir,
  "02 Classical ML"
)

# Classical ML subdirectories
models_dir <- file.path(
  classical_dir,
  "models"
)

results_classical_dir <- file.path(
  classical_dir,
  "results"
)

figures_classical_dir <- file.path(
  classical_dir,
  "figures"
)

# QSVC folder
qsvc_dir <- file.path(
  project_dir,
  "03 Quantum_ML",
  "05 QSVC"
)

# QSVC data folder
data_dir <- file.path(
  qsvc_dir,
  "data"
)

# Create QSVC data folder if needed
dir.create(
  data_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

# VERIFY PATHS
# ==========================================================

cat("Project directory:\n")
print(project_dir)

cat("\nClassical ML directory:\n")
print(classical_dir)

cat("\nModels directory:\n")
print(models_dir)

cat("\nQSVC directory:\n")
print(qsvc_dir)

cat("\nQSVC data directory:\n")
print(data_dir)

# ==========================================================
# VERIFY SHARED TRAIN-TEST SPLIT
# ==========================================================

train_index_path <- file.path(
  models_dir,
  "BinaryClass_Train_Index.rds"
)

cat("\nFINAL SHARED SPLIT PATH:\n")
print(train_index_path)
cat("\nShared split exists: ")
print(file.exists(train_index_path))

cat("\nLooking for:\n")
print(train_index_path)

if (!file.exists(train_index_path)) {

  stop(
    paste0(
      "\nERROR: Shared train-test split not found.\n\n",
      "Expected location:\n",
      train_index_path
    )
  )

}

cat(
  "\nSUCCESS: BinaryClass_Train_Index.rds found.\n"
)



cat(
  "\nShared BinaryClass_Train_Index.rds found successfully.\n"
)

# ==========================================================
# 6. LOAD THE EXACT SHARED TRAIN-TEST SPLIT
# ==========================================================

train_index <- readRDS(train_index_path)

cat("\nShared train-test split loaded successfully.\n")

cat("\nNumber of training indices:\n")
print(length(train_index))

# ==========================================================
# 7. LOAD THE ZELLER MICROBIOME DATASET
# ==========================================================

zeller <- curatedMetagenomicData(
  "ZellerG_2014.relative_abundance",
  dryrun = FALSE
)[[1]]

# ==========================================================
# 8. EXTRACT ABUNDANCE MATRIX
# ==========================================================

abundance <- assay(zeller)

# Original structure:
# rows    = microbial features
# columns = samples
#
# Transpose so:
# rows    = samples
# columns = microbial features

X <- as.data.frame(
  t(abundance)
)

# ==========================================================
# 9. EXTRACT DISEASE LABELS
# ==========================================================

y <- colData(zeller)$study_condition

# ==========================================================
# 10. CREATE WORKING DATASET
# ==========================================================

crc_data <- X

crc_data$disease <- y

# ==========================================================
# 11. CREATE THE FINAL BINARY DATASET
# ==========================================================

binary_data <- crc_data %>%
  
  mutate(
    disease = case_match(
      disease,
      "control" ~ "healthy",
      .default = disease
    )
  ) %>%
  
  filter(
    disease %in% c(
      "healthy",
      "CRC"
    )
  )

# ==========================================================
# 12. SET THE EXACT CLASS ORDER
# ==========================================================

binary_data$disease <- factor(
  binary_data$disease,
  levels = c(
    "healthy",
    "CRC"
  )
)

# ==========================================================
# 13. VERIFY THE BINARY DATASET
# ==========================================================

cat("\nBinary dataset dimensions:\n")
print(dim(binary_data))

cat("\nBinary class counts:\n")
print(table(binary_data$disease))

# Expected:
# healthy = 61
# CRC     = 53
# Total   = 114

# ==========================================================
# 14. VERIFY SHARED INDICES MATCH THIS DATASET
# ==========================================================

if (max(train_index) > nrow(binary_data)) {

  stop(
    paste0(
      "\nERROR: The loaded train indices do not match ",
      "the current binary dataset.\n",
      "Maximum train index: ", max(train_index), "\n",
      "Number of binary samples: ", nrow(binary_data)
    )
  )

}

# ==========================================================
# 15. CREATE THE EXACT SHARED TRAIN-TEST SPLIT
# ==========================================================

train_data <- binary_data[
  train_index,
  ,
  drop = FALSE
]

test_data <- binary_data[
  -train_index,
  ,
  drop = FALSE
]

# ==========================================================
# 16. VERIFY SPLIT DIMENSIONS
# ==========================================================

cat("\nTraining dataset dimensions:\n")
print(dim(train_data))

cat("\nTesting dataset dimensions:\n")
print(dim(test_data))

if (nrow(train_data) != 92) {

  stop(
    paste0(
      "\nERROR: Expected 92 training samples but found ",
      nrow(train_data)
    )
  )

}

if (nrow(test_data) != 22) {

  stop(
    paste0(
      "\nERROR: Expected 22 testing samples but found ",
      nrow(test_data)
    )
  )

}

cat(
  "\nCorrect shared split confirmed:",
  "\nTraining samples = 92",
  "\nTesting samples  = 22\n"
)

# ==========================================================
# 17. VERIFY CLASS DISTRIBUTION
# ==========================================================

cat("\nTraining class distribution:\n")
print(table(train_data$disease))

cat("\nTesting class distribution:\n")
print(table(test_data$disease))

cat("\nTraining class proportions:\n")
print(prop.table(table(train_data$disease)))

cat("\nTesting class proportions:\n")
print(prop.table(table(test_data$disease)))

# ==========================================================
# 18. SEPARATE FEATURES AND LABELS
# ==========================================================

X_train_raw <- train_data %>%
  select(-disease)

X_test_raw <- test_data %>%
  select(-disease)

y_train <- train_data$disease

y_test <- test_data$disease

# ==========================================================
# 19. NEAR-ZERO VARIANCE FILTERING
# Training data only
# ==========================================================

nzv <- nearZeroVar(
  X_train_raw
)

cat("\nNumber of NZV features removed:\n")
print(length(nzv))

if (length(nzv) > 0) {

  X_train_nzv <- X_train_raw[
    ,
    -nzv,
    drop = FALSE
  ]

  X_test_nzv <- X_test_raw[
    ,
    -nzv,
    drop = FALSE
  ]

} else {

  X_train_nzv <- X_train_raw

  X_test_nzv <- X_test_raw

}

cat("\nFeatures remaining after NZV filtering:\n")
print(ncol(X_train_nzv))

# ==========================================================
# 20. BORUTA FEATURE SELECTION
# Training data only
# ==========================================================

set.seed(1910)

boruta_data <- data.frame(
  Disease = y_train,
  X_train_nzv,
  check.names = FALSE
)

boruta_model <- Boruta(
  Disease ~ .,
  data = boruta_data,
  doTrace = 0
)

boruta_final <- TentativeRoughFix(
  boruta_model
)

selected_features <- getSelectedAttributes(
  boruta_final,
  withTentative = FALSE
)

# ==========================================================
# Rank the 20 Boruta-confirmed features by Boruta importance
# ==========================================================

library(tibble)

boruta_stats <- attStats(boruta_final) %>%
  as.data.frame() %>%
  rownames_to_column("Original") %>%
  filter(
    Original %in% selected_features,
    decision == "Confirmed"
  ) %>%
  arrange(desc(meanImp))

# Verify that all 20 selected features are present
stopifnot(nrow(boruta_stats) == 20)

# Keep the top 8 features for the QSVC
top8_boruta <- boruta_stats %>%
  slice_head(n = 8) %>%
  mutate(Rank = row_number()) %>%
  select(
    Rank,
    Original,
    meanImp,
    medianImp,
    minImp,
    normHits,
    decision
  )

print(top8_boruta)
# ==========================================================
# 21. VERIFY EXACTLY 20 FEATURES WERE SELECTED
# ==========================================================

cat("\nNumber of Boruta-selected features:\n")
print(length(selected_features))

if (length(selected_features) != 20) {

  stop(
    paste0(
      "\nERROR: Expected exactly 20 Boruta-selected features, ",
      "but found ",
      length(selected_features),
      ".\n",
      "Do not continue because this would break consistency ",
      "with the final classical pipeline."
    )
  )

}

cat(
  "\nExactly 20 Boruta-selected features confirmed.\n"
)

# ==========================================================
# 22. DISPLAY THE SELECTED FEATURES
# ==========================================================

cat("\nSelected features:\n")

print(
  selected_features
)

# ==========================================================
# 23. SUBSET TRAINING DATA
# ==========================================================

X_train_boruta <- X_train_nzv[
  ,
  selected_features,
  drop = FALSE
]

# ==========================================================
# 24. SUBSET TESTING DATA
# Using the exact same training-selected features
# ==========================================================

X_test_boruta <- X_test_nzv[
  ,
  selected_features,
  drop = FALSE
]

# ==========================================================
# 25. VERIFY FEATURE CONSISTENCY
# ==========================================================

if (!identical(
  colnames(X_train_boruta),
  colnames(X_test_boruta)
)) {

  stop(
    "\nERROR: Training and testing feature names/order do not match."
  )

}

cat(
  "\nTraining and testing feature order confirmed identical.\n"
)

# ==========================================================
# 26. CENTER AND SCALE
# Fit on training data only
# ==========================================================

scaler <- preProcess(
  X_train_boruta,
  method = c(
    "center",
    "scale"
  )
)

# ==========================================================
# 27. TRANSFORM TRAINING DATA
# ==========================================================

X_train_scaled <- predict(
  scaler,
  X_train_boruta
)

# ==========================================================
# 28. TRANSFORM TEST DATA
# Using the training-fitted scaler
# ==========================================================

X_test_scaled <- predict(
  scaler,
  X_test_boruta
)

cat(
  "\nTraining and testing data scaled successfully.\n"
)

# ==========================================================
# 29. CREATE FEATURE NAME MAP
# ==========================================================

feature_name_map <- data.frame(
  
  Original = colnames(X_train_scaled),
  
  Syntactic = make.names(
    colnames(X_train_scaled),
    unique = TRUE
  ),
  
  stringsAsFactors = FALSE
  
)

# ==========================================================
# Match Boruta-ranked features to their Python-safe names
# ==========================================================

top8_export <- top8_boruta %>%
  left_join(
    feature_name_map,
    by = "Original"
  ) %>%
  select(
    Rank,
    Original,
    Syntactic,
    meanImp,
    medianImp,
    minImp,
    normHits,
    decision
  )

# Verify every top-8 feature was successfully mapped
stopifnot(nrow(top8_export) == 8)
stopifnot(!anyNA(top8_export$Syntactic))

print(top8_export)

# ==========================================================
# 30. APPLY PYTHON-SAFE FEATURE NAMES
# ==========================================================

colnames(X_train_scaled) <- feature_name_map$Syntactic

colnames(X_test_scaled) <- feature_name_map$Syntactic

# ==========================================================
# 31. VERIFY FINAL FEATURE ORDER
# ==========================================================

if (!identical(
  colnames(X_train_scaled),
  colnames(X_test_scaled)
)) {

  stop(
    "\nERROR: Final training and testing feature order does not match."
  )

}

cat(
  "\nFinal feature names and order confirmed identical.\n"
)

# ==========================================================
# 32. PRESERVE ORIGINAL SAMPLE IDS
# ==========================================================

train_sample_ids <- rownames(
  X_train_scaled
)

test_sample_ids <- rownames(
  X_test_scaled
)

# ==========================================================
# 33. CHECK FOR TRAIN-TEST SAMPLE OVERLAP
# ==========================================================

overlap_ids <- intersect(
  train_sample_ids,
  test_sample_ids
)

if (length(overlap_ids) > 0) {

  stop(
    "\nERROR: Sample IDs overlap between training and testing data."
  )

}

cat(
  "\nNo training/testing sample overlap detected.\n"
)

# ==========================================================
# 34. CREATE TRAINING EXPORT DATAFRAME
# ==========================================================

train_export <- data.frame(
  
  SampleID = train_sample_ids,
  
  X_train_scaled,
  
  disease = y_train,
  
  check.names = FALSE
  
)

# ==========================================================
# 35. CREATE TESTING EXPORT DATAFRAME
# ==========================================================

test_export <- data.frame(
  
  SampleID = test_sample_ids,
  
  X_test_scaled,
  
  disease = y_test,
  
  check.names = FALSE
  
)

# ==========================================================
# 36. VERIFY FINAL EXPORT DIMENSIONS
# ==========================================================

cat("\nTraining export dimensions:\n")
print(dim(train_export))

cat("\nTesting export dimensions:\n")
print(dim(test_export))

# Expected:
# 92 rows × 22 columns
# = SampleID + 20 features + disease
#
# 22 rows × 22 columns
# = SampleID + 20 features + disease

if (nrow(train_export) != 92) {
  stop("\nERROR: Training export does not contain 92 samples.")
}

if (nrow(test_export) != 22) {
  stop("\nERROR: Testing export does not contain 22 samples.")
}

if (ncol(train_export) != 22) {
  stop(
    paste0(
      "\nERROR: Expected 22 columns in training export ",
      "(SampleID + 20 features + disease), but found ",
      ncol(train_export)
    )
  )
}

if (ncol(test_export) != 22) {
  stop(
    paste0(
      "\nERROR: Expected 22 columns in testing export ",
      "(SampleID + 20 features + disease), but found ",
      ncol(test_export)
    )
  )
}

# ==========================================================
# 37. EXPORT TRAINING DATA
# ==========================================================

train_export_path <- file.path(
  data_dir,
  "QSVC_Train_Data.csv"
)

write.csv(
  train_export,
  train_export_path,
  row.names = FALSE
)

# ==========================================================
# 38. EXPORT TESTING DATA
# ==========================================================

test_export_path <- file.path(
  data_dir,
  "QSVC_Test_Data.csv"
)

write.csv(
  test_export,
  test_export_path,
  row.names = FALSE
)

# ==========================================================
# 39. EXPORT FEATURE NAME MAP
# ==========================================================

feature_map_path <- file.path(
  data_dir,
  "QSVC_Feature_Name_Map.csv"
)

write.csv(
  feature_name_map,
  feature_map_path,
  row.names = FALSE
)

# ==========================================================
# 40. EXPORT TOP 8 FEATURES USED FOR QSVC
# ==========================================================

write.csv(
  top8_export,
  file.path(data_dir, "QSVC_Top8_Features.csv"),
  row.names = FALSE
)

# ==========================================================
# 41. CREATE EXPORT LOG
# ==========================================================

log_path <- file.path(
  data_dir,
  "QSVC_Export_Log.txt"
)

log_lines <- c(

  "==========================================================",
  "QSVC DATA EXPORT LOG",
  "==========================================================",
  
  paste(
    "Export date:",
    format(Sys.time())
  ),
  
  paste(
    "Random seed:",
    1910
  ),
  
  "",
  
  "CLASSIFICATION TASK",
  "Healthy vs CRC",
  
  "",
  
  "DATASET",
  "ZellerG_2014.relative_abundance",
  
  "",
  
  "SHARED TRAIN-TEST SPLIT",
  paste(
    "Split file:",
    train_index_path
  ),
  
  paste(
    "Training samples:",
    nrow(train_export)
  ),
  
  paste(
    "Testing samples:",
    nrow(test_export)
  ),
  
  "",
  
  "FEATURE SELECTION",
  
  paste(
    "NZV features removed:",
    length(nzv)
  ),
  
  paste(
    "Final Boruta-selected features:",
    length(selected_features)
  ),
  
  "",
  
  "TRAINING CLASS COUNTS",
  
  paste(
    names(table(y_train)),
    table(y_train),
    sep = " = ",
    collapse = ", "
  ),
  
  "",
  
  "TESTING CLASS COUNTS",
  
  paste(
    names(table(y_test)),
    table(y_test),
    sep = " = ",
    collapse = ", "
  ),
  
  "",
  
  "OUTPUT FILES",
  
  train_export_path,
  test_export_path,
  feature_map_path
  
)

writeLines(
  log_lines,
  log_path
)

# ==========================================================
# 41. FINAL VERIFICATION
# ==========================================================

cat(
  "\n==========================================================\n"
)

cat(
  "QSVC DATA EXPORT COMPLETED SUCCESSFULLY\n"
)

cat(
  "==========================================================\n"
)

cat(
  "\nOutput directory:\n",
  data_dir,
  "\n"
)

cat(
  "\nFiles created:\n"
)

cat(
  "\n1. ",
  train_export_path
)

cat(
  "\n2. ",
  test_export_path
)

cat(
  "\n3. ",
  feature_map_path
)

cat(
  "\n4. ",
  log_path
)

cat(
  "\n\nFinal summary:\n"
)

cat(
  "Training samples: ",
  nrow(train_export),
  "\n"
)

cat(
  "Testing samples: ",
  nrow(test_export),
  "\n"
)

cat(
  "Final features: ",
  length(selected_features),
  "\n"
)

cat(
  "Seed: 1910\n"
)

cat(
  "\nReady for Python QSVC analysis.\n"
)
```

## Session information

The final section records the exact R session used for this knit. The same information is also saved as a text artifact so the HTML report and repository contain both a human-readable and machine-readable provenance record.

```{r session-information}
session_file <- file.path(
  results_dir,
  "session_info",
  "01_QSVC_Data_Preparation_sessionInfo.txt"
)
write_session_info(session_file)
cat("Saved session information to:", session_file, "\n\n")
sessionInfo()
```
