---
title: "QSVC External Validation Prerequisite Export"
author: "Harshini"
date: "2026-08-25"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "##",
  fig.width = 7.2,
  fig.height = 5.2,
  dpi = 300,
  fig.align = "center",
  out.width = "92%",
  results = "markup"
)
```

# Purpose

This R Markdown file prepares the two independent external cohorts for the **existing discovery-trained QSVC pipeline**.

It must **not** perform feature selection, refit the discovery scaler, retrain any model, or modify the existing QSVC notebook.

The frozen discovery artifacts are:

- `02 Classical ML/results/Binary_Selected_20_Features.csv`
- `02 Classical ML/models/Scaler.rds`
- `03 Quantum_ML/05 QSVC/data/QSVC_Feature_Name_Map.csv`

The resulting external CSV files are written only to:

`03 Quantum_ML/05 QSVC/data/`

The exported external matrices contain the same 20 discovery-selected microbial features, after applying the frozen discovery center/scale transformation. Python then applies the already-fitted 8-feature quantum angle scaler.

```{r packages}
# ==========================================================
# 1. Packages
# ==========================================================

library(curatedMetagenomicData)
library(TreeSummarizedExperiment)
library(SummarizedExperiment)
library(caret)
library(dplyr)
library(tibble)

set.seed(1910)
```

```{r paths}
# ==========================================================
# 2. Project paths
# ==========================================================

# This Rmd lives in: project_root/03 Quantum_ML/05 QSVC/
# Knit uses the Rmd directory as the working directory, so go up
# two levels to recover the repository root.
project_dir <- normalizePath(
  file.path(getwd(), "../.."),
  winslash = "/",
  mustWork = TRUE
)

# Guard against accidentally knitting from the wrong directory.
if (!dir.exists(file.path(project_dir, "02 Classical ML")) ||
    !dir.exists(file.path(project_dir, "03 Quantum_ML"))) {
  stop(
    "Could not identify the CRC project root from the QSVC notebook location.
",
    "Expected both '02 Classical ML' and '03 Quantum_ML' under: ",
    project_dir
  )
}

classical_dir <- file.path(project_dir, "02 Classical ML")
qsvc_dir <- file.path(project_dir, "03 Quantum_ML", "05 QSVC")

models_dir <- file.path(classical_dir, "models")
classical_results_dir <- file.path(classical_dir, "results")
qsvc_data_dir <- file.path(qsvc_dir, "data")

dir.create(qsvc_data_dir, recursive = TRUE, showWarnings = FALSE)

cat("Project:", project_dir, "\n")
cat("QSVC data directory:", qsvc_data_dir, "\n")
```

```{r frozen-artifact-check}
# ==========================================================
# 3. Verify frozen discovery artifacts
# ==========================================================

features_path <- file.path(
  classical_results_dir,
  "Binary_Selected_20_Features.csv"
)

scaler_path <- file.path(
  models_dir,
  "Scaler.rds"
)

qsvc_feature_map_path <- file.path(
  qsvc_data_dir,
  "QSVC_Feature_Name_Map.csv"
)

required_files <- c(
  features_path,
  scaler_path,
  qsvc_feature_map_path
)

cat("\nFrozen artifact paths being checked:\n")
cat("Features :", features_path, "\n")
cat("Scaler   :", scaler_path, "\n")
cat("Feature map:", qsvc_feature_map_path, "\n\n")

missing_files <- required_files[!file.exists(required_files)]

if (length(missing_files) > 0) {
  stop(
    "Required frozen discovery artifact(s) missing:\n",
    paste(missing_files, collapse = "\n")
  )
}

selected_20 <- read.csv(
  features_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
)$Feature

scaler <- readRDS(scaler_path)

qsvc_map <- read.csv(
  qsvc_feature_map_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

stopifnot(length(selected_20) == 20)
stopifnot(all(selected_20 %in% qsvc_map$Original))

cat("Frozen 20-feature panel verified.\n")
cat("Frozen scaler verified.\n")
cat("QSVC feature-name map verified.\n")
```

```{r export-function}
# ==========================================================
# 4. External-cohort export function
# ==========================================================

export_external_cohort <- function(
  cohort_name,
  output_filename
) {

  cat("\n====================================================\n")
  cat("Processing:", cohort_name, "\n")
  cat("====================================================\n")

  cmd_obj <- curatedMetagenomicData(
    paste0(cohort_name, ".relative_abundance"),
    dryrun = FALSE
  )[[1]]

  abundance <- assay(cmd_obj)
  X_raw <- as.data.frame(t(abundance), check.names = FALSE)

  disease_raw <- colData(cmd_obj)$study_condition
  sample_ids <- colnames(cmd_obj)

  raw_df <- X_raw
  raw_df$SampleID <- sample_ids
  raw_df$disease_raw <- as.character(disease_raw)

  raw_df$disease <- dplyr::case_match(
    raw_df$disease_raw,
    "control" ~ "healthy",
    .default = raw_df$disease_raw
  )

  # Binary Healthy-vs-CRC task only.
  binary_df <- raw_df %>%
    filter(disease %in% c("healthy", "CRC"))

  binary_df$disease <- factor(
    binary_df$disease,
    levels = c("healthy", "CRC")
  )

  # Require unique sample IDs.
  if (anyDuplicated(binary_df$SampleID) > 0) {
    stop("Duplicate sample IDs detected in ", cohort_name)
  }

  # Require all frozen discovery features.
  missing_features <- setdiff(
    selected_20,
    colnames(binary_df)
  )

  if (length(missing_features) > 0) {
    stop(
      cohort_name,
      " is missing one or more frozen discovery features:\n",
      paste(missing_features, collapse = "\n")
    )
  }

  X_ext <- binary_df[, selected_20, drop = FALSE]

  # Apply the discovery-trained scaler only.
  X_ext_scaled <- predict(
    scaler,
    X_ext
  )

  # Convert to the exact syntactic names used by QSVC_Train_Data.csv.
  syntactic_names <- qsvc_map$Syntactic[
    match(
      colnames(X_ext_scaled),
      qsvc_map$Original
    )
  ]

  if (anyNA(syntactic_names)) {
    stop(
      "Failed to map one or more frozen features to QSVC syntactic names in ",
      cohort_name
    )
  }

  colnames(X_ext_scaled) <- syntactic_names

  # Preserve the feature order exactly.
  if (!identical(
    colnames(X_ext_scaled),
    qsvc_map$Syntactic[match(selected_20, qsvc_map$Original)]
  )) {
    stop(
      "External feature order does not match the frozen QSVC feature order in ",
      cohort_name
    )
  }

  export_df <- data.frame(
    SampleID = binary_df$SampleID,
    X_ext_scaled,
    disease = as.character(binary_df$disease),
    check.names = FALSE
  )

  # Structural checks.
  stopifnot(nrow(export_df) > 0)
  stopifnot(ncol(export_df) == 22)
  stopifnot(length(setdiff(c("healthy", "CRC"), unique(export_df$disease))) == 0)
  # Check missing values on the scaled feature matrix itself.
# At this point the feature columns have been renamed to the
# exact QSVC syntactic names, so selected_20 (the original names)
# must not be used to index export_df here.
stopifnot(!anyNA(X_ext_scaled))

  output_path <- file.path(
    qsvc_data_dir,
    output_filename
  )

 write.csv(
  export_df,
  output_path,
  row.names = FALSE
)

  summary <- data.frame(
    Cohort = cohort_name,
    N_Total = nrow(export_df),
    N_Healthy = sum(export_df$disease == "healthy"),
    N_CRC = sum(export_df$disease == "CRC"),
    Feature_Count = 20,
    Scaler = "Frozen Zeller discovery Scaler.rds",
    Feature_Source = "Frozen Binary_Selected_20_Features.csv",
    Output_File = output_path,
    stringsAsFactors = FALSE
  )

  summary
}
```

```{r export-fengq}
# ==========================================================
# 5. Export FengQ_2015
# ==========================================================

fengq_summary <- export_external_cohort(
  cohort_name = "FengQ_2015",
  output_filename = "QSVC_External_FengQ_2015_Data.csv"
)

fengq_summary
```

```{r export-vogtmann}
# ==========================================================
# 6. Export VogtmannE_2016
# ==========================================================

vogtmann_summary <- export_external_cohort(
  cohort_name = "VogtmannE_2016",
  output_filename = "QSVC_External_VogtmannE_2016_Data.csv"
)

vogtmann_summary
```

```{r combined-summary}
# ==========================================================
# 7. Combined summary and export log
# ==========================================================

external_summary <- rbind(
  fengq_summary,
  vogtmann_summary
)

summary_path <- file.path(
  qsvc_data_dir,
  "QSVC_External_Validation_Cohort_Summary.csv"
)

write.csv(
  external_summary,
  summary_path,
  row.names = FALSE
)

log_lines <- c(
  "QSVC EXTERNAL VALIDATION PREREQUISITE EXPORT",
  "=============================================",
  paste("Export date:", format(Sys.time())),
  paste("Random seed:", 1910),
  "",
  "FROZEN DISCOVERY ARTIFACTS",
  features_path,
  scaler_path,
  qsvc_feature_map_path,
  "",
  "EXTERNAL OUTPUTS",
  file.path(qsvc_data_dir, "QSVC_External_FengQ_2015_Data.csv"),
  file.path(qsvc_data_dir, "QSVC_External_VogtmannE_2016_Data.csv"),
  summary_path,
  "",
  "IMPORTANT",
  "No external feature selection was performed.",
  "No external scaler was fitted.",
  "No model was retrained.",
  "The exported features are the frozen discovery 20-feature panel",
  "after applying the frozen discovery-trained center/scale transformation."
)

writeLines(
  log_lines,
  file.path(
    qsvc_data_dir,
    "QSVC_External_Validation_Prerequisite_Log.txt"
  )
)

print(external_summary)
cat("\nPrerequisite export completed successfully.\n")
cat("Outputs written to:\n", qsvc_data_dir, "\n")
```
